---
title: "LLMs for corpus annotation: Close and extended replications of Morin & Marttinen Larsson (2025) using smaller open LLMs"
author: Author1 and Author2
date: last-modified
language: 
    title-block-author-single: "Authors"
execute: 
  warning: false
format:
  html:
    number-sections: true
  pdf:
    number-sections: true
bibliography: references.bib
csl: unified-style-sheet-for-linguistics.csl
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| label: close_replication_results
#| include: false

library(tidyverse)
library(irr)
library(caret)
library(knitr)

comparison_all <- read_csv2(
  "replication_data/comparison_all_results.csv",
  show_col_types = FALSE
)

# Test distribution

test_distribution <- table(
  comparison_all$Human_classification
)

majority_count <- max(test_distribution)
test_total <- nrow(comparison_all)
majority_accuracy <- majority_count / test_total

# Accuracy
gpt_accuracy <- mean(
  comparison_all$Human_classification == comparison_all$GPT_classification
)

mistral_accuracy <- mean(
  comparison_all$Human_classification == comparison_all$Mistral_classification
)

claude_accuracy <- mean(
  comparison_all$Human_classification == comparison_all$Claude_classification
)

# Classification order for confusion matrices
classification_levels <- c(
  "evaluative",
  "non-evaluative"
)

# Confusion matrices
actual_mistral <- factor(
  comparison_all$Human_classification,
  levels = classification_levels
)

predicted_mistral <- factor(
  comparison_all$Mistral_classification,
  levels = classification_levels
)

conf_matrix_mistral <- confusionMatrix(
  data = predicted_mistral,
  reference = actual_mistral,
  positive = "evaluative"
)


actual_GPT_close <- factor(
  comparison_all$Human_classification,
  levels = classification_levels
)

predicted_GPT_close <- factor(
  comparison_all$GPT_classification,
  levels = classification_levels
)

conf_matrix_GPT_close <- confusionMatrix(
  data = predicted_GPT_close,
  reference = actual_GPT_close,
  positive = "evaluative"
)


actual_claude <- factor(
  comparison_all$Human_classification,
  levels = classification_levels
)

predicted_claude <- factor(
  comparison_all$Claude_classification,
  levels = classification_levels
)

conf_matrix_claude <- confusionMatrix(
  data = predicted_claude,
  reference = actual_claude,
  positive = "evaluative"
)


# Calculate Matthews correlation coefficient
calculate_mcc <- function(cm) {
  
  TP <- cm[1, 1]
  FN <- cm[2, 1]
  FP <- cm[1, 2]
  TN <- cm[2, 2]
  
  denominator <- sqrt(
    (TP + FP) *
      (TP + FN) *
      (TN + FP) *
      (TN + FN)
  )
  
  if (denominator == 0) {
    return(NA_real_)
  }
  
  (TP * TN - FP * FN) / denominator
}


# Create table with metrics for all three models
metrics_table <- data.frame(
  
  Metric = c(
    "Accuracy",
    "Precision",
    "Recall",
    "F1-score",
    "Matthews correlation coefficient"
  ),

  Mistral = c(
    as.numeric(conf_matrix_mistral$overall["Accuracy"]),
    as.numeric(conf_matrix_mistral$byClass["Pos Pred Value"]),
    as.numeric(conf_matrix_mistral$byClass["Sensitivity"]),
    as.numeric(conf_matrix_mistral$byClass["F1"]),
    calculate_mcc(conf_matrix_mistral$table)
  ),
  
  GPT = c(
    as.numeric(conf_matrix_GPT_close$overall["Accuracy"]),
    as.numeric(conf_matrix_GPT_close$byClass["Pos Pred Value"]),
    as.numeric(conf_matrix_GPT_close$byClass["Sensitivity"]),
    as.numeric(conf_matrix_GPT_close$byClass["F1"]),
    calculate_mcc(conf_matrix_GPT_close$table)
  ),
  
  Claude = c(
    as.numeric(conf_matrix_claude$overall["Accuracy"]),
    as.numeric(conf_matrix_claude$byClass["Pos Pred Value"]),
    as.numeric(conf_matrix_claude$byClass["Sensitivity"]),
    as.numeric(conf_matrix_claude$byClass["F1"]),
    calculate_mcc(conf_matrix_claude$table)
  )
)

# Format Accuracy as whole percentage
metrics_table[1, -1] <- paste0(
  round(as.numeric(metrics_table[1, -1]) * 100, 0),
  "%"
)

# Format Precision, Recall and F1-score (as percentages with one decimal place)
metrics_table[2:4, -1] <- lapply(
  metrics_table[2:4, -1],
  function(x) {
    paste0(round(as.numeric(x) * 100, 1), "%")
  }
)

# Format MCC to two decimal places (not percent)
metrics_table[5, -1] <- lapply(
  metrics_table[5, -1],
  function(x) {
    sprintf("%.2f", as.numeric(x))
  }
)
```

```{r}
#| label: reduced-training-2
#| include: false

gpt_reduced_2 <- read_csv2(
  "reduced_training/GPT_reduced_training_2_results.csv",
  show_col_types = FALSE
)

# Accuracy
gpt_reduced_2_accuracy <- mean(
  gpt_reduced_2$Human_classification ==
    gpt_reduced_2$GPT_classification
)

# Number of agreements
gpt_reduced_2_agreement <- sum(
  gpt_reduced_2$Human_classification ==
    gpt_reduced_2$GPT_classification
)

# Number of test items
gpt_reduced_2_total <- nrow(gpt_reduced_2)

# Confusion matrices
actual_gpt_reduced_2 <- factor(
  gpt_reduced_2$Human_classification,
  levels = classification_levels
)

predicted_gpt_reduced_2 <- factor(
  gpt_reduced_2$GPT_classification,
  levels = classification_levels
)

conf_matrix_gpt_reduced_2 <- confusionMatrix(
  data = predicted_gpt_reduced_2,
  reference = actual_gpt_reduced_2,
  positive = "evaluative"
)

# Calculate Matthews correlation coefficient
calculate_mcc <- function(cm) {
  
  # Extract the four cells from confusion matrix
  TP <- cm[1, 1]
  FN <- cm[2, 1]
  FP <- cm[1, 2]
  TN <- cm[2, 2]
  
  # Calculate denominator
  denominator <- sqrt(
    (TP + FP) *
      (TP + FN) *
      (TN + FP) *
      (TN + FN)
  )
  
  # Avoid division by 0
  if (denominator == 0) {
    return(NA_real_)
  }
  
  # Calculate Matthews correlation coefficient
  (TP * TN - FP * FN) / denominator
}


# Create table with metrics for all three models
metrics_table_ext <- data.frame(
  
  # Names of metrics
  Metric = c(
    "Accuracy",
    "Precision",
    "Recall",
    "F1-score",
    "Matthews correlation coefficient"
  ),
  
  # GPT close run
  `GPT (close replication)` = c(
    as.numeric(conf_matrix_GPT_close$overall["Accuracy"]),
    as.numeric(conf_matrix_GPT_close$byClass["Pos Pred Value"]),
    as.numeric(conf_matrix_GPT_close$byClass["Sensitivity"]),
    as.numeric(conf_matrix_GPT_close$byClass["F1"]),
    calculate_mcc(conf_matrix_GPT_close$table)
  ),
  
  # GPT reduced training
  `GPT (reduced training)` = c(
    as.numeric(conf_matrix_gpt_reduced_2$overall["Accuracy"]),
    as.numeric(conf_matrix_gpt_reduced_2$byClass["Pos Pred Value"]),
    as.numeric(conf_matrix_gpt_reduced_2$byClass["Sensitivity"]),
    as.numeric(conf_matrix_gpt_reduced_2$byClass["F1"]),
    calculate_mcc(conf_matrix_gpt_reduced_2$table)
  ),
  
  # Claude
  `Claude (original study)` = c(
    as.numeric(conf_matrix_claude$overall["Accuracy"]),
    as.numeric(conf_matrix_claude$byClass["Pos Pred Value"]),
    as.numeric(conf_matrix_claude$byClass["Sensitivity"]),
    as.numeric(conf_matrix_claude$byClass["F1"]),
    calculate_mcc(conf_matrix_claude$table)
  )
)


# Format Accuracy as whole percentage
metrics_table_ext[1, -1] <- paste0(
  round(as.numeric(metrics_table_ext[1, -1]) * 100, 0),
  "%"
)


# Format Precision, Recall and F1-score
# as percentages with one decimal place
metrics_table_ext[2:4, -1] <- lapply(
  metrics_table_ext[2:4, -1],
  function(x) {
    paste0(
      round(as.numeric(x) * 100, 1),
      "%"
    )
  }
)


# Format MCC to two decimal places
metrics_table_ext[5, -1] <- lapply(
  metrics_table_ext[5, -1],
  function(x) {
    sprintf(
      "%.2f",
      as.numeric(x)
    )
  }
)
```

# Abstract {.unnumbered}

Large language models (LLMs) are increasingly being leveraged for corpus
annotation. In an early evaluation of such an application, *Linguistics
Vanguard* published "a replicable method for automating grammatical
annotation" [@morin2025]. Using Claude 3.5 Sonnet, it involved several
'conversational' phases of training and testing. The study reported an
accuracy of 93% and concluded that the pipeline may also be effective
for other linguistic tasks, including in languages other than English.
We first aimed to replicate this "replicable method" as closely as
possible, using the same data, but with two open LLMs. While a 24B
Mistral model did not perform as well (67% accuracy), a 120B open GPT
model yielded equally good results (94% accuracy). This suggest that, as
of 2026, large commercial LLMs may not be necessary for this type of
annotation task. In a second replication study, we tested the need for
the extensive 'in-conversation' training process outlined in @morin2025.
We compared it to a much simpler few-shot training approach, concluding
that the training process need not be as elaborate as in the original
study. Our results have important implications for the replicability of
linguistic research involving LLMs. Although the authors published their
data and prompts, we found that crucial information necessary to
replicate the procedure was missing. We argue that issues around
accessibility, reproducibility, data leakage, and carbon footprint can
and should be taken into consideration when choosing a tool and warn
against the risk of overgeneralising the results of a single, binary
classification task in English to other linguistic tasks and languages.

**Keywords**: Replication, Large language model, Reproducibility, Corpus
linguistics, Open Science.

<!--# ELF: The submission guidelines state "Your article should include an abstract of approx. 200 words (please note, citations to sources are not permitted in the abstract) and 3-5 relevant keywords". I don't think it makes sense to remove the citation of the original study but I have removed page references and have also reduced the length of the abstract considerably. Although it's still too long I suggest we leave like this for the first submission. -->

# Introduction {#sec-Intro}

Over the past decades, linguists have gained access to ever larger
corpora of natural language. This poses great opportunities, but also
greatly increases the workload associated with the manual and/or
(semi-)automated preprocessing and annotation of large amounts of data.
The rise of large language models (LLMs) is therefore an exciting
development for linguistics research, offering new possibilities for
complementing and extending existing automated methods for processing
larger datasets and corpora. In recent years, a growing number of
publications have reported on the usefulness of LLMs for a range of
linguistic annotation tasks, such as the annotation of speech acts
[@yu2024; @su2025; @ostyakova2025], metaphor detection [@fuoli2026],
word sense or function disambiguation [@fonteyn2024; @morin2025], and
political text classification and annotation [@heseltine2024].

While LLMs are increasingly prevalent in both linguistic research and
our everyday lives, they also raise a number of concerns. For research
purposes, specifically, their 'black-box' nature raises concerns over
the transparency, reproducibility, replicability of the research
produced with their assistance. Note, that although the terms have been
used interchangeably in the past, we distinguish between
*reproducibility* and *replicability* following the definitions of the
FORRT [@parsonsCommunitysourcedGlossaryOpen2022] and Turing Way
[@theturingwaycommunityTuringWayHandbook2022] communities. We thus
define *reproducibility* as the ability of an independent researcher or
team to obtain the same results as an original study using the original
data and methods. This is in contrast to *replicability*, where the same
methods are applied, but on different data; and robustness, where the
same data, but different methods are used (see
@fig-ReproducibilityTerminology). Only if an effect can be reliably
observed across different datasets with different methods, can we claim
that it is *generalisable*.

![Term disambiguation [@theturingwaycommunityTuringWayHandbook2022, CC
BY
4.0]](figures/TuringWay_reproducible-matrix.jpg){#fig-ReproducibilityTerminology
out-width="70%"
fig-alt="A table defining four terms: reproducible, replicable, robust, and generalisable. For reproducibility, data and analysis are the same. For replicability, data is different and the analysis is the same. For robustness, data is the same and the analysis is different. For generalisability, data and analysis are both different."
width="60%"}

In this study, we seek to replicate @morin2025's methodological study,
whose title claims to present a "replicable method" for leveraging LLMs
for linguistic annotation. The original study was conducted using a
commercial, closed-source LLM: Anthropic's model Claude 3.5 Sonnet
[@morin2025: 4]. It is no longer possible to reproduce the original
results as Claude Sonnet 3.5, which was released in June 2024, was
retired in October 2025
(<https://web.archive.org/web/20260911184540/https://killedbyai.net/dead/claude-3-5-sonnet/>),
just 18 months after its release and only six months after the online
version of the original article was published. We therefore sought to
apply the original methodological pipeline to the same data, but using
two open-source LLMs, whose accessibility can be guaranteed for
independent verification and future replications and whose parameters
can be controlled to generate reproducible results (see @sec-CloseRep).
In an extended replication, we subsequently re-ran the classification
task with a much simpler pipeline than that proposed in the original
paper (see @sec-ExtRep).

In the meantime, the authors of the original study have published a
follow-up paper as a preprint [@morin2026], in which they further
develop their "large-scale pipeline for automatic corpus annotation
using LLMs" based on the same case study of the English *consider*
construction. Although there are many similarities in its aims, this
follow-up study adopts considerably simpler prompting than @morin2025
and the case study involves a larger dataset. It also pursues a somewhat
different focus, placing greater emphasis on the different types of
evaluative uses of the *consider* construction and on the role of genre.
However, the classification tasks are again conducted using a large,
commercial model -- this time from OpenAI [@morin2026: 17]. As such, we
believe that evaluating the reproducibility of the original pipeline
with smaller, open LLMs remains a very relevant and worthwhile
undertaking.

While large, cloud-based, proprietary LLMs continue to top performance
benchmarks, open-source LLMs have made significant progress in closing
this performance gap [@manchanda: 4]. We consider open models to form a
crucial part of the open research ecosystem (alongside open materials,
open data, open code, open documentation, open access, etc.). Openness
promotes transparency, reproducibility, and quality control; which are
necessary features to support accessible, reliable, and credible
empirical findings [@liesenfeld2023]. In the discussion and conclusion
section of their subsequent study, @morin2026 detail the limitations of
using closed-source, commercial LLMs for linguistic research; however,
both their studies do not test the necessity of relying on these
products to achieve high accuracy. This is what we set out to do in the
present study.

# Research aims and questions {#sec-RQs}

The present study presents close and extended replication studies of the
results published in:

> Morin, Cameron, and Matti Marttinen Larsson. 2025. “Large Corpora and
> Large Language Models: A Replicable Method for Automating Grammatical
> Annotation.” *Linguistics Vanguard* 11 (1): 501–10.
> <https://doi/10.1515/lingvan-2024-0228>.

We are grateful to the original authors who published the data and
materials associated with this project in an OSF repository
[@morin2024]. The repository contains a "conversation transcript", from
which the iterative training process could be traced. It also contains
several datasets for training and testing an LLM to differentiate
between evaluative and non-evaluative uses of the verb *consider*, which
can be illustrated with the following examples:

a.  Cain is *considered* a master of American hard boiled crime fiction
    (…) [@morin2025: 2]

b.  Please *consider* all classes when punching number in to
    spread-sheet and designing new items [@morin2025: 2].

In example a. the verb *consider* constitutes an evaluation of the
subject. Such evaluative constructions can be and are often formulated
using *consider as* or *consider to be*. In example b., however, the
meaning is equivalent to 'take into account' or 'think about', and has
no evaluative component.

The model outputs classified sentences as 'evaluative', 'non-evaluative'
or '?' in ambiguous cases. They were evaluated against a human gold
standard also provided in @morin2024.

In this study, we seek answers to the following three research questions
(RQs):

1.  Is the "replicable method" proposed in @morin2025 independently
    replicable using the information and materials made available by the
    authors in the article and associated OSF repository?

2.  Following the original pipeline, can smaller, open-source models
    perform as well or better on @morin2025's *consider* annotation task
    as Claude Sonnet 3.5 based on the accuracy metrics reported in
    @morin2025\[: 6\]?

3.  If RQ2 is answered in the positive, can this also be achieved with a
    significantly less extensive training process than that proposed in
    @morin2025?

While the close replication seeks answers to RQ1 and RQ2, our extended
replication serves to answer RQ3. We test the performance of two
open-source LLMs: Mistral Small 3.2 24B Instruct 2506 [@mistralai2026]
(hereafter referred to as 'Mistral 24B'), an open-weight model released
in 2025 under Apache 2.0 license by the French AI company Mistral AI,
and OpenAI GPT OSS 120B [@openai2026], an open-weight large language
model that was also released under Apache 2.0 license by OpenAI
(hereafter 'GPT 120B').

The outcome of the replications will be compared and assessed based on
the metrics @morin2025 provided in their evaluation: Accuracy,
Precision, Recall, F1 score, and Matthew's correlation coefficient
[@morin2025: 6]. We formulate the following hypotheses:

Hypothesis 1: The model OpenAI GPT OSS 120B performs as well as or
better than Claude Sonnet 3.5 on the *consider* task, as measured by the
accuracy metrics reported in @morin2025.

Hypothesis 2: The model Mistral Small 3.2 24B Instruct 2506 performs as
well as or better than Claude Sonnet 3.5 on the *consider* task, as
measured by the accuracy metrics reported in @morin2025.

Hypothesis 3: The annotation task can be performed with a significantly
less extensive training process than that used by @morin2025, while
achieving performance comparable to or better than Claude Sonnet 3.5
according to the accuracy metrics reported in @morin2025.

<!--# ELF: I've had to remove the bullet points for the hypotheses as otherwise RStudio refused to view the doc in visual mode. As you will have noticed, I have also added a RQ and renumbered the hypotheses. -->

# Close replication {#sec-CloseRep}

@morin2025 present a methodological pipeline leveraging an LLM for the
automatic annotation of grammatical data, based on the case study of the
evaluative *consider* construction in English. Having engaged in a
complex "conversation" with Claude 3.5 Sonnet over several iterative
phases of training and testing (see @fig-Pipeline), @morin2026\[: 6\]
were able to achieve an accuracy of 93% compared to the gold standard
provided by a human annotator.

![Morin & Marttinen Larsson (2025)’s pipeline for LLM-assisted
annotation of linguistic data, based on the case study of consider
constructions [CC BY @morin2025:
5]](figures/MorinLarsson2026-10.png){#fig-Pipeline fig-align="center"}

@tbl-CloseRepAcc compares the original accuracy metrics with those of
our close replication in which we attempted to follow the "replicable
method" as closely as possible, using the same data and matching the
training and testing to a conversation transcript the authors provided
[@morin2024], but with two open LLMs. As explained in @sec-Intro, the
commercial LLM used in the original study, Claude Sonnet 3.5, was
already deprecated by October 2025, just six months after the online
publication of the study. Whilst the authors could not have predicted
that Anthropic would retire this particular model so quickly, this
extremely quick turnaround illustrates the risks of attempting to build
reproducible and sustainable research pipelines on the commercial
services of AI companies.

In answer to RQ1, when designing our close replication, we found that
essential information to independently replicate the annotation task was
missing, forcing us to make some educated guesses as to the exact
procedure. Most importantly, the original paper does not disclose how
the pipeline was implemented at the programmatic level. This includes
whether the web interface or an API was used[^1], how the model
parameters were configured, which programming language was used, and how
the datasets were provided as input to the models. Since the article
only provides an outline of the iterative training approach, many
details of the actual process had to be inferred from the so-called
"conversation transcript" provided in the OSF repository, which contains
all the "answers" and explanations given to the model during the
training phases. It was unclear to us whether the authors used Claude's
web interface or an API when we designed our replication studies, but we
chose to use an API and to share all of our scripts to ensure that our
replications are as reproducible as possible (see Data Availability
Statement). Whilst not essential, having access to the authors' original
analysis code would also have been useful to ensure more precise
comparability of our results.

[^1]: While this information was missing in @morin2025, the authors
    disclosed that they used Anthrophic's web chat interface in a more
    recent preprint [@morin2026]. We designed and conducted our
    replication studies prior to this publication and were not aware of
    this follow-up study until recently.

```{r, echo=FALSE}
#| label: tbl-CloseRepAcc

kable(
  metrics_table,
  col.names = c(
    "Metric",
    "Mistral 3.2 24B",
    "GPT OSS 120B",
    "Claude 3.5 Sonnet"
  ),
  caption = "Classification metrics of the three models compared to the human gold standard"
)
```

Based on accuracy compared to the original human gold standard, our open
120B GPT model showed comparable and even slightly better performance
than Claude in the original study (GPT: `r round(100 * gpt_accuracy)`%
vs. Claude `r round(100 * claude_accuracy)`%), while 24B Mistral
performed considerably worse (`r round(100 * mistral_accuracy)`%). This
observation is reflected in the other metrics as well: While the GPT
model's performance is similar or better in all evaluated metrics,
Mistral's much smaller model fails to achieve comparable results. While
Mistral's 67% accuracy might initially sound moderately good,
investigating these results as 2 x 2 matrices (see `0_Appendix.pdf`)
reveals that it is heavily biased toward predicting 'non-evaluative'
meaning. This poor accuracy is best observed in the extremely low recall
of
`r sprintf("%.1f%%", as.numeric(conf_matrix_mistral$byClass["Sensitivity"]) * 100)`.
Given the distribution of *consider*-meanings in the testing dataset
('non-evaluative' was the majority class accounting for
`r majority_count`/`r test_total`, `r round(100 * majority_accuracy)`%
of examples), a classifier that always predicted the majority class
would also have achieved an accuracy of
`r round(100 * majority_accuracy)`%. The imbalance in this dataset is
common to many linguistic annotation tasks and serves as a reminder
that, as did @morin2025\[Table 1\], it is crucial to consider a range of
metrics -- not just accuracy -- when assessing model performance.

Our results suggest that the replicability of @morin2025's approach
using open LLMs may depend on the size of the models. In answer to RQ2,
the open GPT model with 120 B parameters did perform at least as well
than Claude 3.5 Sonnet, which is thought to have had around 175B
parameters [@abacha2024: 6], whereas the considerably smaller 24B
Mistral model did not. We therefore have strong evidence in support of
hypothesis 1, but can reject hypothesis 2.

# Extended replication {#sec-ExtRep}

Our close replication raised questions about the necessity of using very
large, commercial LLMs to complete this type of binary linguistic
classification task in standard English. With the present extended
replication, we turn to testing the need for the extensive
'in-conversation' training and evaluation process proposed in @morin2025
(see also @fig-Pipeline). To this end, we applied a much simpler,
few-shot training approach using the best-performing open LLM from our
first replication, GPT OSS 120B, and the original data.

As shown in @tbl-ExtRepAcc, only instructing the GPT model using the
pretraining set and then directly proceeding to testing, resulted in an
accuracy of `r round(100 * gpt_reduced_2_accuracy)`%
(`r gpt_reduced_2_agreement`/`r gpt_reduced_2_total`). While the overall
accuracy is equivalent to the open GPT result in our close replication,
recall is even higher at 100%, meaning that this simpler process led to
the GPT model not missing any positives ('evaluative' sentences).
Compared to the original study, the accuracy metrics of this extended
replication are also comparable: only precision is slightly lower (GPT:
`r round(conf_matrix_gpt_reduced_2$byClass["Pos Pred Value"] * 100, 1)`%
vs. Claude:
`r round(conf_matrix_claude$byClass["Pos Pred Value"] * 100, 1)`%),
meaning that the open GPT model predicted marginally more false
positives ('evaluative' for actual 'non-evaluative') than Claude Sonnet
3.5 using the considerably more complex 'in-conversation' training
procedure. The other accuracy metrics are higher than in the original
study. These results provide clear support for our third hypothesis.

```{r, echo=FALSE}
#| label: tbl-ExtRepAcc
kable(
  metrics_table_ext,
  col.names = c(
    "Metric",
    "GPT (close replication)",
    "GPT (reduced training)",
    "Claude (original study)"
  ),
  caption = "Classification metrics of both GPT analyses and Claude compared to the human gold standard"
)
```

# Discussion

Our results suggest that, when using a comparable open-source LLM,
similarly high accuracy rates can be achieved with a considerably less
elaborate "training" process than that described in @morin2025. The
original study claimed that "the systematic prompt engineering and
training steps in our approach help mitigate potential biases in
Claude’s training data" [@morin2025: 7]. Interestingly, in a recent
paper following up on this project [@morin2026], the authors themselves
propose a much simpler workflow with which they reach equally high
accuracy rates on the same and a more complex classification task of the
*consider* construction.

As mentioned in @sec-Intro, open models constitute an integral part of
Open Science practices, which also include open data, code, and
materials, pre-registration and registered reports, open access
publishing, and aim to make research more accessible, transparent,
reproducible, replicable, and ultimately more robust and reliable [see
e.g. @parsonsCommunitysourcedGlossaryOpen2022]. In the following, we
reflect more generally on the replicability of corpus-processing
pipelines that involve LLMs focussing on the physical and intellectual
accessibility, computational reproducibility and generalisability, as
well as ethical and environmental concerns associated with the use of
both closed and open LLMs [which, as @liesenfeld2023l explain, is by no
means not a clear-cut binary distinction].

## Accessibility

In the context of using LLMs for corpus annotation, several dimensions
of accessibility must be considered. We begin with the physical
accessibility of the models themselves.

A primary factor influencing physical accessibility is financial cost.
The cost of commercial LLMs is highly variable — ranging from free tiers
to expensive enterprise plans — and can change unpredictably. For
instance, in their follow-up study, @morin2026 describe switching from
Anthropic’s Claude models to OpenAI’s GPT-5, citing Anthropic’s
substantially higher API fees as the reason. This transition allowed
them to complete two classification tasks across over 140,000
concordance lines for a total of \$104 USD [@morin2026: 18]. However,
this figure does not account for the significant costs incurred during
their initial, more elaborate evaluation of the Claude Sonnet 3.5 model
in @morin2025. Running open models, however, can also requires
substantial equipment or infrastructure. For example, @shakir2026's
comparison of 11 small to mid-size open LLMs for corpus annotation was
completed locally on a laptop equipped with an RTX 4090 GPU and 24 GB of
RAM, a set-up that carries a considerable price tag. For the present
project, such hardware was unavailable; we were constrained to a
considerably less powerful laptop, which limited our choice of open
models to those accessible via API through our university’s platform. In
this light, some "free" or "cheap" commercial models may appear more
physically accessible than open ones, as they do not require local
installation or server management. Yet, this apparent advantage is
offset by a long-term financial vulnerability inherent to proprietary
systems. If a commercial provider alters its pricing, discontinues a
service, or restricts API access, as has already occurred in some cases,
a previously accessible method may suddenly no longer be useable.

The second dimension of accessibility is intellectual. A major argument
in favour of commercial LLMs put forward by @morin2025 and others is
their availability through web-based chatbots, which require no coding
skills. Indeed, @morin2025 [: 7] justified their choice of a Claude
model by noting that "it provides a powerful and accessible chat-based
interface that does not require programming expertise," thereby making
their method "accessible to a maximally wide usership" [@morin2025: 3].
However, their latest publication appears to mark a shift away from this
original motivation. The revised pipeline described in @morin2026
involves (admittedly relatively basic) Python code to interact with
OpenAI’s API and to process data in batches. While this introduces a
programming requirement, the authors provide their code, which they note
"can be adapted to different APIs and research contexts" [@morin2025:
14]. This sharing of code significantly enhances the accessibility of
their revised approach, even if it raises the technical bar for
replication.

## Reproducibility

Reproducibility is a cornerstone of scientific inquiry, yet it is also a
matter of research sustainability. If the results of a study cannot be
reproduced, its findings remain provisional, and the resources invested
in producing them may yield only limited long-term value. In the context
of LLM-based corpus annotation, achieving reproducibility is fraught
with challenges that extend beyond the usual difficulties of documenting
methodological detail. As the present reproduction study has shown, even
with the best of intentions, providing sufficient information to allow
for independent replication is more difficult than might be expected.

A fundamental obstacle is the inherent non-determinism of commercial
cloud-based LLMs. Even when hyperparameters such as temperature can be
set and are documented, the outputs of these models are never fully
reproducible. This is due to a combination of factors, including
server-side load balancing, model updates that occur without notice, and
the fact that the exact model version, system prompt, and other aspects
of a model's internal state are usually opaque to the user. This
inherent variability undermines the very notion of computational
reproducibility, regardless of how meticulously the prompts and
procedures are documented. The original study by @morin2025 exemplifies
several additional limitations that further constrain replicability.
First, the specific Claude model used is no longer available. Second,
important implementation details were not explicitly reported. In
particular, the fact that the study relied on Claude's chat-based
interface rather than the API is only explicitly stated in the follow-up
publication [@morin2026].

The use of a chat interface is particularly problematic for
reproducibility, as it introduces additional layers of variability:
interfaces may apply their own formatting, truncate, or even modify
prompts without notifying the user, and they rarely allow for the
systematic control of parameters that APIs typically afford. This lack
of transparency makes it difficult for other researchers to determine
whether differences between an original study and its replication are
attributable to the models themselves, to implementation choices, or to
other unaccounted factors.

A final reproducibility concern relates to the human gold standard used
for evaluation. The original study appears to have relied on annotations
from a single annotator, and it is unclear whether inter-rater agreement
was assessed during the creation of this gold standard. Without such
validation, the reliability of the evaluation itself is uncertain, which
in turn complicates the interpretation of any replication attempt. If
the gold standard is idiosyncratic or contains errors, then high
accuracy scores may reflect overfitting to a particular annotator's
judgments rather than genuine model competence.

Taken together, these issues highlight that reproducibility in LLM-based
research is not merely a matter of good documentation, but also of
choosing tools and methods that are inherently more amenable to
replication. Open models, when run locally with fixed versions and
documented hyperparameters, offer a greater degree of control and
transparency. However, as we discuss in the next subsection, even
locally-run open models present their own challenges with respect to
generalisability and environmental impact.

## Generalisability 

In their conclusion section, @morin2025\[: 7\] claim to have "put
forward this annotation method as a replicable one, which linguists can
use for the analysis of other grammatical constructions, in English and
in other languages". While the authors also acknowledge that "individual
parameters, instructions, and evaluation criteria are of course bound to
vary depending on the specific construction under study" and that some
languages will be harder to study due to less available training data
\[@morin2025: 7\], the generalisability of LLM-based annotation
pipelines to other tasks, languages, and contexts warrants careful
consideration.

A primary issue is the linguistic bias inherent to the majority of LLMs,
including those made available by AI companies as 'open' models such as
the OpenAI GPT OSS model used in the present reproduction. Research so
far has predominantly focused on English-language contexts, with
considerably less attention paid to LLMs trained on or evaluated in
other languages [@yao2025: 2; @kargaran]. Models trained on
English-centric corpora often struggle with underrepresented languages
due to limited multilingual reasoning training data, weaker language
processing abilities, and lower-quality 'reasoning chains'. Moreover,
latent 'reasoning' processes, i.e. those that occur internally within
the model, are still almost exclusively examined in English [@liu2026:
22330]. This linguistic skew has direct implications for corpus
annotation: a pipeline that performs well on English data may yield
substantially less accurate results when applied to other languages,
particularly those with limited digital presence.

A further concern relates to the intransparency of commercial LLM
development. Closed models are developed with no public oversight or
ethical scrutiny [@liesenfeld2023]. Their training relies on vast
amounts of data, including text from the internet, publicly available
datasets, and proprietary sources. This opacity raises two distinct
problems. First, these models are susceptible to passive privacy
leakage, whereby sensitive information contained in training data may be
revealed or reused in generation. Second, they are vulnerable to various
forms of active privacy attacks [@yan2025: 1]. Both issues are
particularly acute for linguistic research, where corpora may contain
personal or sensitive data.

A related and often overlooked problem is the risk of data leakage in
replication studies. Most so-called 'open' LLMs are in fact only
open-weight models and are developed by companies that do not disclose
their training data [@liesenfeld2023]. This means that the training data
of such models could conceivably include the very test sets that were
publicly released by the original study. If this were the case, high
accuracy scores on a replication task might reflect the model having
memorised the evaluation data rather than demonstrating genuine
generalisation ability. This issue is particularly insidious because it
is nearly impossible to detect without access to the training data,
which is typically not provided. Given these challenges, we argue that
it will always be necessary to conduct thorough evaluations of
annotation procedures before deploying them in a linguistic study. This
is particularly true for under-resourced languages and for more complex
or less documented tasks, where the risk of poor performance is higher
and the consequences of inaccurate annotation are more severe.
Researchers should therefore treat LLM-based annotation pipelines with
caution, validating their outputs against human annotations and
carefully considering the linguistic and contextual scope of their
claims.

## Ethical and environmental concerns

Beyond the practical considerations of accessibility, reproducibility,
and generalisability, the use of LLMs in research raises significant
ethical and environmental questions that also merit careful reflection.
These concerns apply to both closed and open models, albeit in different
ways, and should be weighed before deciding whether an LLM-based
approach is appropriate for a given task.

We cannot ignore the substantial environmental cost associated with
LLMs. Training and fine-tuning models with billions of parameters
requires vast computational resources, including large amounts of
electricity and water for cooling data centres, which in turn generate
significant carbon emissions [@özcan; @kron]. Importantly, the
environmental impact of LLMs is not limited to the training phase; it
also depends on the complexity and length of individual prompts, with
more demanding prompts generally requiring more computational resources
than shorter and less complex ones [@jegham2025: 11-12]. Although AI
systems are becoming cheaper and faster, their global adoption is
rapidly increasing overall resource consumption [@jegham2025]. This
creates a paradox: efficiency gains are offset by exponential growth in
usage. Given these costs, we argue that researchers should critically
assess whether an LLM is the most appropriate tool for their specific
task. In addition, @shakir2026 have shown that smaller, locally-run
models can perform as well as, or in some cases even outperform, larger,
more computationally intensive ones. These finding suggests that the
default assumption of "bigger is better" is not always warranted, and
that researchers should consider more resource-efficient alternatives
where possible. Whilst calculating the environmental footprint LLM usage
is always tricky, it is notoriously difficult when accessing cloud-based
models as providers rarely disclose energy consumption or carbon
emissions for individual requests. This opacity makes it challenging for
researchers to make fully informed decisions about the environmental
trade-offs involved.

Yet environmental impact is only one of several ethical considerations
that warrant attention. Depending on the nature of the annotation task,
corporate or governmental interests could be insidiously influencing the
results of LLM-based annotations. This is particularly relevant when
working on politically sensitive topics such as gender, where model
outputs may reflect the biases or agendas of the companies that
developed them. Most so-called 'frontier' models, whether closed or
(partially) open, are currently developed by American and Chinese
companies, whose priorities may not align with the values of academic
research. The use of such models raises serious questions about research
integrity and academic sovereignty: to what extent should linguists rely
on tools whose development is driven by commercial or geopolitical
interests, and whose training data and internal mechanisms are largely
opaque?

In sum, we argue that the decision to use an LLM in a research pipeline
should not be made solely on the basis of accuracy or accessibility.
Researchers must also consider the environmental costs, the potential
for biased or agenda-driven outputs, and the broader implications for
academic sovereignty. While we acknowledged that these factors are
difficult to quantify in practice, we believe that they are no less
important for being hard to measure.

# Conclusion

This study investigated the replicability of an LLM-based methodological
pipeline for linguistic data annotation. While the smaller 24B model did
not achieve comparable results, @morin2025's approach was successfully
replicated with the larger 120B open-weight model, as well as with a
significantly less computationally intensive approach. At the same time,
it is worth noting that the study by @morin2025 represents an early
contribution to this specific area. Regarding the generalisability of
this method, it is important to consider that the grammatical
construction annotated here constitutes a rather clear-cut formal case
[@morin2025: 7]. Even with promising results from some LLMs, further
exploration is needed to effectively and reliably leverage them for
linguistic data annotation at large, and we strongly caution against the
risk of overgeneralising from this relatively simple binary
classification task in standard English to more complex tasks in
less-resourced languages.

The extensions explored in this study give insights into what can be
learned from testing LLM-based methods beyond the conditions of the
original study. They illustrate the importance of investigating not only
which results an approach produces, but also under which conditions it
remains reliable and replicable. As LLMs are increasingly used in
corpus-based research, open-science practices, such as making datasets,
code, model parameters, and evaluation procedures as transparent as
possible, are more relevant than ever.

More broadly, this study highlights the importance of replicability in
linguistic research with LLMs and the advantages of using (more) open
models, which can offer at least some amount of control, and greater
possibilities for reproducibility. To leverage LLMs in linguistic
research, it is important to weigh their potential and their
limitations. This also means considering broader issues associated with
their usage, ranging from data leakage to environmental impact.

## Data availability statement {.unnumbered}

The data and code that support the findings of this study are openly
available in \[repository name e.g “figshare”\] at
<http://doi.org/>\[doi\], reference number \[reference number\]. The
current repository has been anonymised for double-blind peer review, but
will be de-anonymised on acceptance.

## CRediT author statement {.unnumbered}

**Author1**: Methodology, Data Curation, Code, Validation, Formal
analysis, Investigation, Writing - Original Draft, Writing - Review &
Editing, Project administration. **Author2**: Conceptualisation,
Methodology, Validation, Investigation, Writing - Original Draft,
Writing - Review & Editing, Supervision, Project administration.

## Declaration of conflict of interest {.unnumbered}

The authors have no conflicts of interest to declare.

# References
