---
title: "Appendix"
format: html
editor: visual
execute:
  echo: false
---

# Full results

```{r}
#| label: setup
#| include: false
#| warning: false
#| message: false

library(readr)
library(irr)
library(knitr)
library(caret)
library(kableExtra)
```

```{r}
#| label: appendix-results-close-replication
#| include: false

comparison_all <- read_csv2(
  "replication_data/comparison_all_results.csv",
  show_col_types = FALSE
)

# Test set distribution
test_distribution <- table(
  comparison_all$Human_classification
)

majority_count <- max(test_distribution)
test_total <- nrow(comparison_all)
majority_accuracy <- majority_count / test_total

# Accuracy
gpt_accuracy <- mean(
  comparison_all$Human_classification ==
    comparison_all$GPT_classification
)

mistral_accuracy <- mean(
  comparison_all$Human_classification ==
    comparison_all$Mistral_classification
)

claude_accuracy <- mean(
  comparison_all$Human_classification ==
    comparison_all$Claude_classification
)

# Number of agreements
gpt_agreement <- sum(
  comparison_all$Human_classification ==
    comparison_all$GPT_classification
)

mistral_agreement <- sum(
  comparison_all$Human_classification ==
    comparison_all$Mistral_classification
)

claude_agreement <- sum(
  comparison_all$Human_classification ==
    comparison_all$Claude_classification
)

# Cohen's kappa
gpt_kappa <- kappa2(
  comparison_all[c("Human_classification", "GPT_classification")]
)$value

mistral_kappa <- kappa2(
  comparison_all[c("Human_classification", "Mistral_classification")]
)$value

claude_kappa <- kappa2(
  comparison_all[c("Human_classification", "Claude_classification")]
)$value

# Classification order for confusion matrices
classification_levels <- c(
  "evaluative",
  "non-evaluative"
)

```

### Distribution of classifications in the testing set

```{r}
#| label: tbl-test-distribution
#| tbl-cap: "Distribution of classifications in the test set according to the human gold standard"

test_distribution_table <- data.frame(
  Classification = names(test_distribution),
  n = as.vector(test_distribution),
  Percentage = paste0(
    round(100 * as.vector(test_distribution) / test_total),
    "%"
  )
)

test_distribution_table <- rbind(
  test_distribution_table,
  data.frame(
    Classification = "Total",
    n = test_total,
    Percentage = "100%"
  )
)

kable(test_distribution_table)
```

## Close replication

The following tables represent the results of the close replication.

### Accuracy and Cohen's Kappa compared to the human gold standard

```{r}
#| label: close-rep-tbl-full-results
#| tbl-cap: "Classification accuracy and agreement with the human gold standard"

results_table <- data.frame(
  Model = c("Claude", "GPT", "Mistral"),
  Accuracy = c(
    paste0(round(100 * claude_accuracy), "%"),
    paste0(round(100 * gpt_accuracy), "%"),
    paste0(round(100 * mistral_accuracy), "%")
  ),
  Alignments = c(
    paste0(claude_agreement, "/", nrow(comparison_all)),
    paste0(gpt_agreement, "/", nrow(comparison_all)),
    paste0(mistral_agreement, "/", nrow(comparison_all))
  ),
  `Cohen's κ` = round(
    c(claude_kappa, gpt_kappa, mistral_kappa),
    2
  )
)

kable(results_table)

```

### Tables representing accuracy as 2 x 2 matrices

```{r}
#| label: tbl-gpt-close-confusion
#| tbl-cap: "GPT classifications compared with the human gold standard"
#| warning: false
#| message: false

actual_GPT_close <- factor(
  comparison_all$Human_classification,
  levels = classification_levels
)

predicted_GPT_close <- factor(
  comparison_all$GPT_classification,
  levels = classification_levels
)

conf_matrix_GPT_close <- confusionMatrix(
  data = predicted_GPT_close,
  reference = actual_GPT_close,
  positive = "evaluative"
)

#print(conf_matrix_GPT_close$table)
#kable(conf_matrix_GPT_close$table)

# create table output with model names
kable(
  conf_matrix_GPT_close$table,
  caption = "Prediction (GPT) vs Reference (Human)")

```

```{r}
#| label: tbl-mistral-confusion
#| tbl-cap: "Mistral classifications compared with the human gold standard"
#| warning: false
#| message: false

actual_mistral <- factor(
  comparison_all$Human_classification,
  levels = classification_levels
)

predicted_mistral <- factor(
  comparison_all$Mistral_classification,
  levels = classification_levels
)

conf_matrix_mistral <- confusionMatrix(
  data = predicted_mistral,
  reference = actual_mistral,
  positive = "evaluative"
)

kable(
  conf_matrix_mistral$table,
  caption = "Prediction (Mistral) vs Reference (Human)")
```

```{r}
#| label: tbl-claude-confusion
#| tbl-cap: "Claude classifications compared with the human gold standard"
#| warning: false
#| message: false

actual_claude <- factor(
  comparison_all$Human_classification,
  levels = classification_levels
)

predicted_claude <- factor(
  comparison_all$Claude_classification,
  levels = classification_levels
)

conf_matrix_claude <- confusionMatrix(
  data = predicted_claude,
  reference = actual_claude,
  positive = "evaluative"
)

kable(
  conf_matrix_claude$table,
  caption = "Prediction (Claude) vs Reference (Human)")
```

For the Mistral model, the matrix reveals that it is biased toward predicting non-evaluative: it gets 61 of the 63 human non-evaluative cases right, but misses 32 of the 39 evaluative cases.

```{r}
#| label: tbl-accuracy-metrics-close
#| tbl-cap: "Classification performance of the three models compared with the human gold standard"
#| warning: false
#| message: false

# Calculate Matthews correlation coefficient
calculate_mcc <- function(cm) {
  
  # Extract the four cells from confusion matrix
  TP <- cm[1, 1]
  FN <- cm[2, 1]
  FP <- cm[1, 2]
  TN <- cm[2, 2]
  
  # Calculate denominator
  denominator <- sqrt(
    (TP + FP) *
      (TP + FN) *
      (TN + FP) *
      (TN + FN)
  )
  
  # Avoid division by 0
  if (denominator == 0) {
    return(NA_real_)
  }
  
  # Calculate Matthews correlation coefficient
  (TP * TN - FP * FN) / denominator
}


# Create table with metrics for all three models
metrics_table <- data.frame(
  
  # Names of metrics
  Metric = c(
    "Accuracy",
    "Precision",
    "Recall",
    "F1-score",
    "Matthews correlation coefficient"
  ),
  
  # Mistral metrics
  Mistral = c(
    as.numeric(conf_matrix_mistral$overall["Accuracy"]),
    as.numeric(conf_matrix_mistral$byClass["Pos Pred Value"]),
    as.numeric(conf_matrix_mistral$byClass["Sensitivity"]),
    as.numeric(conf_matrix_mistral$byClass["F1"]),
    calculate_mcc(conf_matrix_mistral$table)
  ),
  
  # GPT metrics
  GPT = c(
    as.numeric(conf_matrix_GPT_close$overall["Accuracy"]),
    as.numeric(conf_matrix_GPT_close$byClass["Pos Pred Value"]),
    as.numeric(conf_matrix_GPT_close$byClass["Sensitivity"]),
    as.numeric(conf_matrix_GPT_close$byClass["F1"]),
    calculate_mcc(conf_matrix_GPT_close$table)
  ),
  
  # Claude metrics
  Claude = c(
    as.numeric(conf_matrix_claude$overall["Accuracy"]),
    as.numeric(conf_matrix_claude$byClass["Pos Pred Value"]),
    as.numeric(conf_matrix_claude$byClass["Sensitivity"]),
    as.numeric(conf_matrix_claude$byClass["F1"]),
    calculate_mcc(conf_matrix_claude$table)
  )
)


# Format Accuracy as whole percentage
metrics_table[1, -1] <- paste0(
  round(as.numeric(metrics_table[1, -1]) * 100, 0),
  "%"
)


# Format Precision, Recall and F1-score (as percentages with one decimal place)
metrics_table[2:4, -1] <- lapply(
  metrics_table[2:4, -1],
  function(x) {
    paste0(
      round(as.numeric(x) * 100, 1),
      "%"
    )
  }
)


# Format MCC to two decimal places (not percent)
metrics_table[5, -1] <- lapply(
  metrics_table[5, -1],
  function(x) {
    sprintf("%.2f", as.numeric(x))
  }
)


# Final table
kable(
  metrics_table,
  col.names = c(
    "Metric",
    "Mistral",
    "GPT",
    "Claude (original study)"
  )
)
```

## Extended replication

### Extended replication with GPT: simpler training process

The GPT part includes only the replication where we went from the pretraining dataset straight to testing.

```{r}
#| label: reduced-training-2-results
#| include: false

library(readr)
library(irr)
library(knitr)

gpt_reduced_2 <- read_csv2(
  "reduced_training/GPT_reduced_training_2_results.csv",
  show_col_types = FALSE
)

# Accuracy
gpt_reduced_2_accuracy <- mean(
  gpt_reduced_2$Human_classification ==
    gpt_reduced_2$GPT_classification
)

# Number of agreements
gpt_reduced_2_agreement <- sum(
  gpt_reduced_2$Human_classification ==
    gpt_reduced_2$GPT_classification
)

# Cohen's kappa
gpt_reduced_2_kappa <- kappa2(
  gpt_reduced_2[
    c("Human_classification", "GPT_classification")
  ]
)$value

# Number of test items
gpt_reduced_2_total <- nrow(gpt_reduced_2)

```

```{r}
#| label: tbl-reduced-training-2
#| tbl-cap: "GPT performance in the reduced training replication (only pretraining dataset and testing)"

reduced_training_2_table <- data.frame(
  Accuracy = paste0(
    round(100 * gpt_reduced_2_accuracy),
    "%"
  ),
  Alignments = paste0(
    gpt_reduced_2_agreement,
    "/",
    gpt_reduced_2_total
  ),
  `Cohen's κ` = round(
    gpt_reduced_2_kappa,
    2
  )
)

kable(reduced_training_2_table)
```

### Extended replication with distilBERT

The distilBERT was once trained with only the pretraining dataset. The associated script is called `BERT_working_script.R` and the results can be found in `BERT_transcripts` folder.

```{r}
#| label: bert-results
#| include: false

library(knitr)

load("BERT_transcripts/BERT_experiment_workspace.RData")

bert_accuracy <- mean(
  preds_test == labels_test
)

bert_agreement <- sum(
  preds_test == labels_test
)

bert_total <- length(labels_test)
```

```{r}
#| label: tbl-bert-results
#| tbl-cap: "distilBERT performance on the test set"

bert_results_table <- data.frame(
  Accuracy = paste0(
    round(100 * bert_accuracy),
    "%"
  ),
  Alignments = paste0(
    bert_agreement,
    "/",
    bert_total
  )
)

kable(bert_results_table)
```

```{r}
#| label: bert-confusion-matrix
#| tbl-cap: "Confusion matrix for the original DistilBERT model"

confusion_matrix_bert <- table(
  Actual = factor(
    labels_test,
    levels = c(0, 1),
    labels = c("evaluative", "non-evaluative")
  ),
  Predicted = factor(
    preds_test,
    levels = c(0, 1),
    labels = c("evaluative", "non-evaluative")
  )
)

kable(confusion_matrix_bert)
```

Also trained with a larger training dataset consisting of the pretraining set and the two unsupervised training sets (script: `BERT_extended_script.R)`:

```{r}
#| label: bert-extended-results
#| include: false

library(knitr)

load("BERT_extended_transcripts/BERT_extended.RData")

bert_extended_accuracy <- mean(
  preds_test_extended == labels_test_extended
)

bert_extended_agreement <- sum(
  preds_test_extended == labels_test_extended
)

bert_extended_total <- length(labels_test_extended)
```

```{r}
#| label: tbl-bert-extended-results
#| tbl-cap: "DistilBERT performance on extended training dataset"

bert_extended_results_table <- data.frame(
  Accuracy = paste0(
    round(100 * bert_extended_accuracy),
    "%"
  ),
  Alignments = paste0(
    bert_extended_agreement,
    "/",
    bert_extended_total
  )
)

kable(bert_extended_results_table)

```

```{r}
#| label: bert-extended-confusion-matrix
#| tbl-cap: "Confusion matrix for the extended DistilBERT model"

confusion_matrix_extended <- table(
  Actual = factor(
    labels_test_extended,
    levels = c(0, 1),
    labels = c("evaluative", "non-evaluative")
  ),
  Predicted = factor(
    preds_test_extended,
    levels = c(0, 1),
    labels = c("evaluative", "non-evaluative")
  )
)

kable(confusion_matrix_extended)
```
