---
title: "Appendix"
format: html
editor: visual
execute:
  echo: false
---

# Full results

```{r}
#| label: appendix-results
#| include: false

library(readr)
library(irr)
library(knitr)

comparison_all <- read_csv2(
  "replication_data/comparison_all_results.csv",
  show_col_types = FALSE
)

# Test set distribution
test_distribution <- table(
  comparison_all$Human_classification
)

majority_count <- max(test_distribution)
test_total <- nrow(comparison_all)
majority_accuracy <- majority_count / test_total

# Accuracy
gpt_accuracy <- mean(
  comparison_all$Human_classification ==
    comparison_all$GPT_classification
)

mistral_accuracy <- mean(
  comparison_all$Human_classification ==
    comparison_all$Mistral_classification
)

claude_accuracy <- mean(
  comparison_all$Human_classification ==
    comparison_all$Claude_classification
)

# Number of agreements
gpt_agreement <- sum(
  comparison_all$Human_classification ==
    comparison_all$GPT_classification
)

mistral_agreement <- sum(
  comparison_all$Human_classification ==
    comparison_all$Mistral_classification
)

claude_agreement <- sum(
  comparison_all$Human_classification ==
    comparison_all$Claude_classification
)

# Cohen's kappa
gpt_kappa <- kappa2(
  comparison_all[c("Human_classification", "GPT_classification")]
)$value

mistral_kappa <- kappa2(
  comparison_all[c("Human_classification", "Mistral_classification")]
)$value

claude_kappa <- kappa2(
  comparison_all[c("Human_classification", "Claude_classification")]
)$value

# Classification order for confusion matrices
classification_levels <- c(
  "evaluative",
  "non-evaluative"
)

```

### Distribution of classifications in the testing set

```{r}
#| label: tbl-test-distribution
#| tbl-cap: "Distribution of classifications in the test set according to the human gold standard"

test_distribution_table <- data.frame(
  Classification = names(test_distribution),
  n = as.vector(test_distribution),
  Percentage = paste0(
    round(100 * as.vector(test_distribution) / test_total),
    "%"
  )
)

test_distribution_table <- rbind(
  test_distribution_table,
  data.frame(
    Classification = "Total",
    n = test_total,
    Percentage = "100%"
  )
)

kable(test_distribution_table)
```

## Close replication

The following tables represent the results of the close replication.

### Accuracy and Cohen's Kappa compared to the human gold standard

```{r}
#| label: close-rep-tbl-full-results
#| tbl-cap: "Classification accuracy and agreement with the human gold standard"

results_table <- data.frame(
  Model = c("Claude", "GPT", "Mistral"),
  Accuracy = c(
    paste0(round(100 * claude_accuracy), "%"),
    paste0(round(100 * gpt_accuracy), "%"),
    paste0(round(100 * mistral_accuracy), "%")
  ),
  Alignments = c(
    paste0(claude_agreement, "/", nrow(comparison_all)),
    paste0(gpt_agreement, "/", nrow(comparison_all)),
    paste0(mistral_agreement, "/", nrow(comparison_all))
  ),
  `Cohen's κ` = round(
    c(claude_kappa, gpt_kappa, mistral_kappa),
    2
  )
)

kable(results_table)

```

### Tables representing accuracy as 2 x 2 matrices

```{r}
#| label: tbl-gpt-confusion
#| tbl-cap: "GPT classifications compared with the human gold standard"

gpt_matrix <- table(
  Human = factor(
    comparison_all$Human_classification,
    levels = classification_levels
  ),
  GPT = factor(
    comparison_all$GPT_classification,
    levels = classification_levels
  )
)

kable(gpt_matrix)
```

```{r}
#| label: tbl-mistral-confusion
#| tbl-cap: "Mistral classifications compared with the human gold standard"

mistral_matrix <- table(
  Human = factor(
    comparison_all$Human_classification,
    levels = classification_levels
  ),
  Mistral = factor(
    comparison_all$Mistral_classification,
    levels = classification_levels
  )
)

kable(mistral_matrix)
```

```{r}
#| label: tbl-claude-confusion
#| tbl-cap: "Claude classifications compared with the human gold standard"

claude_matrix <- table(
  Human = factor(
    comparison_all$Human_classification,
    levels = classification_levels
  ),
  Claude = factor(
    comparison_all$Claude_classification,
    levels = classification_levels
  )
)

kable(claude_matrix)
```

For the Mistral model, the matrix reveals that it is biased toward predicting non-evaluative: it gets 61 of the 63 human non-evaluative cases right, but misses 32 of the 39 evaluative cases.

## Extended replication

```{r}
#| label: reduced-training-2-results
#| include: false

library(readr)
library(irr)
library(knitr)

gpt_reduced_2 <- read_csv2(
  "reduced_training/GPT_reduced_training_2_results.csv",
  show_col_types = FALSE
)

# Accuracy
gpt_reduced_2_accuracy <- mean(
  gpt_reduced_2$Human_classification ==
    gpt_reduced_2$GPT_classification
)

# Number of agreements
gpt_reduced_2_agreement <- sum(
  gpt_reduced_2$Human_classification ==
    gpt_reduced_2$GPT_classification
)

# Cohen's kappa
gpt_reduced_2_kappa <- kappa2(
  gpt_reduced_2[
    c("Human_classification", "GPT_classification")
  ]
)$value

# Number of test items
gpt_reduced_2_total <- nrow(gpt_reduced_2)

```

```{r}
#| label: tbl-reduced-training-2
#| tbl-cap: "GPT performance in the reduced-training replication"

reduced_training_2_table <- data.frame(
  Accuracy = paste0(
    round(100 * gpt_reduced_2_accuracy),
    "%"
  ),
  Alignments = paste0(
    gpt_reduced_2_agreement,
    "/",
    gpt_reduced_2_total
  ),
  `Cohen's κ` = round(
    gpt_reduced_2_kappa,
    2
  )
)

kable(reduced_training_2_table)
```

