---
title: "LLMs for corpus annotation: Close and extended replications of Morin & Marttinen Larsson (2025) using smaller open LLMs and BERT"
author: Hannah Boxleitner & Elen Le Foll
format: html
bibliography: references.bib
csl: unified-style-sheet-for-linguistics.csl
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| label: close_replication_results
#| include: false

library(tidyverse)
library(irr)
library(caret)
library(knitr)

comparison_all <- read_csv2(
  "replication_data/comparison_all_results.csv",
  show_col_types = FALSE
)

# Test distribution

test_distribution <- table(
  comparison_all$Human_classification
)

majority_count <- max(test_distribution)
test_total <- nrow(comparison_all)
majority_accuracy <- majority_count / test_total

# Accuracy
gpt_accuracy <- mean(
  comparison_all$Human_classification == comparison_all$GPT_classification
)

mistral_accuracy <- mean(
  comparison_all$Human_classification == comparison_all$Mistral_classification
)

claude_accuracy <- mean(
  comparison_all$Human_classification == comparison_all$Claude_classification
)

# Cohen's kappa
gpt_kappa <- kappa2(
  comparison_all[c("Human_classification", "GPT_classification")]
)$value

mistral_kappa <- kappa2(
  comparison_all[c("Human_classification", "Mistral_classification")]
)$value

claude_kappa <- kappa2(
  comparison_all[c("Human_classification", "Claude_classification")]
)$value

# Classification order for confusion matrices
classification_levels <- c(
  "evaluative",
  "non-evaluative"
)

# Confusion matrices
actual_mistral <- factor(
  comparison_all$Human_classification,
  levels = classification_levels
)

predicted_mistral <- factor(
  comparison_all$Mistral_classification,
  levels = classification_levels
)

conf_matrix_mistral <- confusionMatrix(
  data = predicted_mistral,
  reference = actual_mistral,
  positive = "evaluative"
)


actual_GPT_close <- factor(
  comparison_all$Human_classification,
  levels = classification_levels
)

predicted_GPT_close <- factor(
  comparison_all$GPT_classification,
  levels = classification_levels
)

conf_matrix_GPT_close <- confusionMatrix(
  data = predicted_GPT_close,
  reference = actual_GPT_close,
  positive = "evaluative"
)


actual_claude <- factor(
  comparison_all$Human_classification,
  levels = classification_levels
)

predicted_claude <- factor(
  comparison_all$Claude_classification,
  levels = classification_levels
)

conf_matrix_claude <- confusionMatrix(
  data = predicted_claude,
  reference = actual_claude,
  positive = "evaluative"
)


# Calculate Matthews correlation coefficient
calculate_mcc <- function(cm) {
  
  TP <- cm[1, 1]
  FN <- cm[2, 1]
  FP <- cm[1, 2]
  TN <- cm[2, 2]
  
  denominator <- sqrt(
    (TP + FP) *
      (TP + FN) *
      (TN + FP) *
      (TN + FN)
  )
  
  if (denominator == 0) {
    return(NA_real_)
  }
  
  (TP * TN - FP * FN) / denominator
}


# Create table with metrics for all three models
metrics_table <- data.frame(
  
  Metric = c(
    "Accuracy",
    "Precision",
    "Recall",
    "F1-score",
    "Matthews correlation coefficient"
  ),

  Mistral = c(
    as.numeric(conf_matrix_mistral$overall["Accuracy"]),
    as.numeric(conf_matrix_mistral$byClass["Pos Pred Value"]),
    as.numeric(conf_matrix_mistral$byClass["Sensitivity"]),
    as.numeric(conf_matrix_mistral$byClass["F1"]),
    calculate_mcc(conf_matrix_mistral$table)
  ),
  
  GPT = c(
    as.numeric(conf_matrix_GPT_close$overall["Accuracy"]),
    as.numeric(conf_matrix_GPT_close$byClass["Pos Pred Value"]),
    as.numeric(conf_matrix_GPT_close$byClass["Sensitivity"]),
    as.numeric(conf_matrix_GPT_close$byClass["F1"]),
    calculate_mcc(conf_matrix_GPT_close$table)
  ),
  
  Claude = c(
    as.numeric(conf_matrix_claude$overall["Accuracy"]),
    as.numeric(conf_matrix_claude$byClass["Pos Pred Value"]),
    as.numeric(conf_matrix_claude$byClass["Sensitivity"]),
    as.numeric(conf_matrix_claude$byClass["F1"]),
    calculate_mcc(conf_matrix_claude$table)
  )
)

# Format Accuracy as whole percentage
metrics_table[1, -1] <- paste0(
  round(as.numeric(metrics_table[1, -1]) * 100, 0),
  "%"
)

# Format Precision, Recall and F1-score (as percentages with one decimal place)
metrics_table[2:4, -1] <- lapply(
  metrics_table[2:4, -1],
  function(x) {
    paste0(round(as.numeric(x) * 100, 1), "%")
  }
)

# Format MCC to two decimal places (not percent)
metrics_table[5, -1] <- lapply(
  metrics_table[5, -1],
  function(x) {
    sprintf("%.2f", as.numeric(x))
  }
)
```

```{r}
#| label: reduced-training-2
#| include: false

gpt_reduced_2 <- read_csv2(
  "reduced_training/GPT_reduced_training_2_results.csv",
  show_col_types = FALSE
)

# Accuracy
gpt_reduced_2_accuracy <- mean(
  gpt_reduced_2$Human_classification ==
    gpt_reduced_2$GPT_classification
)

# Number of agreements
gpt_reduced_2_agreement <- sum(
  gpt_reduced_2$Human_classification ==
    gpt_reduced_2$GPT_classification
)

# Cohen's kappa
gpt_reduced_2_kappa <- kappa2(
  gpt_reduced_2[
    c("Human_classification", "GPT_classification")
  ]
)$value

# Number of test items
gpt_reduced_2_total <- nrow(gpt_reduced_2)

# Confusion matrices
actual_gpt_reduced_2 <- factor(
  gpt_reduced_2$Human_classification,
  levels = classification_levels
)

predicted_gpt_reduced_2 <- factor(
  gpt_reduced_2$GPT_classification,
  levels = classification_levels
)

conf_matrix_gpt_reduced_2 <- confusionMatrix(
  data = predicted_gpt_reduced_2,
  reference = actual_gpt_reduced_2,
  positive = "evaluative"
)

# Calculate Matthews correlation coefficient
calculate_mcc <- function(cm) {
  
  # Extract the four cells from confusion matrix
  TP <- cm[1, 1]
  FN <- cm[2, 1]
  FP <- cm[1, 2]
  TN <- cm[2, 2]
  
  # Calculate denominator
  denominator <- sqrt(
    (TP + FP) *
      (TP + FN) *
      (TN + FP) *
      (TN + FN)
  )
  
  # Avoid division by 0
  if (denominator == 0) {
    return(NA_real_)
  }
  
  # Calculate Matthews correlation coefficient
  (TP * TN - FP * FN) / denominator
}


# Create table with metrics for all three models
metrics_table_ext <- data.frame(
  
  # Names of metrics
  Metric = c(
    "Accuracy",
    "Precision",
    "Recall",
    "F1-score",
    "Matthews correlation coefficient"
  ),
  
  # GPT close run
  `GPT (close replication)` = c(
    as.numeric(conf_matrix_GPT_close$overall["Accuracy"]),
    as.numeric(conf_matrix_GPT_close$byClass["Pos Pred Value"]),
    as.numeric(conf_matrix_GPT_close$byClass["Sensitivity"]),
    as.numeric(conf_matrix_GPT_close$byClass["F1"]),
    calculate_mcc(conf_matrix_GPT_close$table)
  ),
  
  # GPT reduced training
  `GPT (reduced training)` = c(
    as.numeric(conf_matrix_gpt_reduced_2$overall["Accuracy"]),
    as.numeric(conf_matrix_gpt_reduced_2$byClass["Pos Pred Value"]),
    as.numeric(conf_matrix_gpt_reduced_2$byClass["Sensitivity"]),
    as.numeric(conf_matrix_gpt_reduced_2$byClass["F1"]),
    calculate_mcc(conf_matrix_gpt_reduced_2$table)
  ),
  
  # Claude
  `Claude (original study)` = c(
    as.numeric(conf_matrix_claude$overall["Accuracy"]),
    as.numeric(conf_matrix_claude$byClass["Pos Pred Value"]),
    as.numeric(conf_matrix_claude$byClass["Sensitivity"]),
    as.numeric(conf_matrix_claude$byClass["F1"]),
    calculate_mcc(conf_matrix_claude$table)
  )
)


# Format Accuracy as whole percentage
metrics_table_ext[1, -1] <- paste0(
  round(as.numeric(metrics_table_ext[1, -1]) * 100, 0),
  "%"
)


# Format Precision, Recall and F1-score
# as percentages with one decimal place
metrics_table_ext[2:4, -1] <- lapply(
  metrics_table_ext[2:4, -1],
  function(x) {
    paste0(
      round(as.numeric(x) * 100, 1),
      "%"
    )
  }
)


# Format MCC to two decimal places
metrics_table_ext[5, -1] <- lapply(
  metrics_table_ext[5, -1],
  function(x) {
    sprintf(
      "%.2f",
      as.numeric(x)
    )
  }
)

```

```{r}
#| label: bert-results
#| include: false

load("BERT_transcripts/BERT_experiment_workspace.RData")

bert_accuracy <- mean(
  preds_test == labels_test
)

bert_agreement <- sum(
  preds_test == labels_test
)

bert_total <- length(labels_test)

# Confusion matrix: first DistilBERT (pretraining only)
actual_bert <- factor(
  labels_test,
  levels = c(0, 1),
  labels = c("evaluative", "non-evaluative")
)

predicted_bert <- factor(
  preds_test,
  levels = c(0, 1),
  labels = c("evaluative", "non-evaluative")
)

conf_matrix_bert <- confusionMatrix(
  data = predicted_bert,
  reference = actual_bert,
  positive = "evaluative"
)

```

```{r}
#| label: bert-extended-results
#| include: false

load("BERT_extended_transcripts/BERT_extended.RData")

bert_extended_accuracy <- mean(
  preds_test_extended == labels_test_extended
)

bert_extended_agreement <- sum(
  preds_test_extended == labels_test_extended
)

bert_extended_total <- length(labels_test_extended)

# Confusion matrix: extended DistilBERT
actual_bert_extended <- factor(
  labels_test_extended,
  levels = c(0, 1),
  labels = c("evaluative", "non-evaluative")
)

predicted_bert_extended <- factor(
  preds_test_extended,
  levels = c(0, 1),
  labels = c("evaluative", "non-evaluative")
)

conf_matrix_bert_extended <- confusionMatrix(
  data = predicted_bert_extended,
  reference = actual_bert_extended,
  positive = "evaluative"
)
```

```{r}
#| label: tbl-accuracy-metrics-bert
#| tbl-cap: "Classification performance of DistilBERT models and Claude compared with the human gold standard"
#| include: false
#| warning: false
#| message: false

metrics_table_bert <- data.frame(
  
  Metric = c(
    "Accuracy",
    "Precision",
    "Recall",
    "F1-score",
    "Matthews correlation coefficient"
  ),
  
  `DistilBERT (pretraining only)` = c(
    unname(conf_matrix_bert$overall["Accuracy"]),
    unname(conf_matrix_bert$byClass["Pos Pred Value"]),
    unname(conf_matrix_bert$byClass["Sensitivity"]),
    unname(conf_matrix_bert$byClass["F1"]),
    calculate_mcc(conf_matrix_bert$table)
  ),
  
  `DistilBERT (extended training)` = c(
    unname(conf_matrix_bert_extended$overall["Accuracy"]),
    unname(conf_matrix_bert_extended$byClass["Pos Pred Value"]),
    unname(conf_matrix_bert_extended$byClass["Sensitivity"]),
    unname(conf_matrix_bert_extended$byClass["F1"]),
    calculate_mcc(conf_matrix_bert_extended$table)
  ),
  
  `Claude (original study)` = c(
    unname(conf_matrix_claude$overall["Accuracy"]),
    unname(conf_matrix_claude$byClass["Pos Pred Value"]),
    unname(conf_matrix_claude$byClass["Sensitivity"]),
    unname(conf_matrix_claude$byClass["F1"]),
    calculate_mcc(conf_matrix_claude$table)
  ),
  
  check.names = FALSE
)


# Format metrics
for (i in 1:nrow(metrics_table_bert)) {
  
  for (j in 2:ncol(metrics_table_bert)) {
    
    value <- as.numeric(metrics_table_bert[i, j])
    
    if (is.na(value) || is.nan(value)) {
      
      metrics_table_bert[i, j] <- "-"
      
    } else if (i == 1) {
      
      metrics_table_bert[i, j] <- paste0(
        round(value * 100, 0),
        "%"
      )
      
    } else if (i %in% 2:4) {
      
      metrics_table_bert[i, j] <- paste0(
        round(value * 100, 1),
        "%"
      )
      
    } else if (i == 5) {
      
      metrics_table_bert[i, j] <- sprintf(
        "%.2f",
        value
      )
    }
  }
}
```

# Abstract

Large Language Models (LLMs) are increasingly being leveraged for the
annotation of large corpora. In an early evaluation of such an
application, @morin2025 presented "a replicable method for automating
grammatical annotation" using Anthropic's model Claude 3.5 Sonnet
[@morin2025: 4]. Their annotation pipeline consisted of a complex
"conversation" with the model over several iterative phases of training
and testing. They were able to achieve an accuracy of 93% compared to
the gold standard provided by a human annotator [@morin2025: 6] and
conclude that the pipeline may also be effective for other linguistic
tasks in other languages [@morin2025: 7].

We aimed to repeat @morin2025's "replicable method" as closely as
possible, using the same data, albeit with two open LLMs, Mistral Small
3.2 24B Instruct [@mistralai2026] and OpenAI GPT OSS 120B [@openai2026].
While the smaller Mistral model did not perform as well on the task
(accuracy of 67%), the open GPT model yielded very similar results
(accuracy of 94%). Although the authors published their data and a full
transcript of their "conversation" with the model [@morin2024], we found
that essential information to independently replicate the study (e.g.,
whether the web interface or an API was used, analysis code, etc.) was
missing - forcing us to make some educated guesses as to the exact
procedure. Our results suggest that, as of early 2026, large commercial
LLMs may not be necessary to complete this kind of linguistic task in
contemporary English.

Next, we tested the need for the extensive 'in-conversation' training
process outlined in @morin2025. To this end, we conducted a second,
extended replication applying a much simpler few-shot training approach
with the best-performing open LLM from our first replication. Our
results suggest that the training process need not be as elaborate as in
@morin2025 to accomplish similarly high accuracy rates.

Finally, we replicated the annotation study using a fine-tuned
DistilBERT model [@sanh]. Although this encoder-only transformer is
considerably smaller than even a small generative LLM, it performed
approximately on par with the 24B Mistral model used in our first
replication. These findings indicate that task-specific supervised
models may provide a competitive and certainly more computationally
efficient alternative to generative LLMs.

Our results have important implications for the replicability of
corpus-linguistic research involving LLMs as part of the annotation
process. We argue that, when choosing a method, issues around model
accessibility, data leakage, and carbon footprint can and should be
taken into consideration.

# Introduction

Over the past decades, linguists have gained access to ever larger
corpora of natural language. This poses great opportunities, but also
considerably increases the workload associated with the manual and/or
(semi-)automated preprocessing and annotation of large amounts of data.
The rise of large language models (LLMs) is therefore an exciting
development for linguistics research, offering new possibilities for
complementing and extending existing automated methods for processing
larger datasets and corpora. In recent years, a growing number of
publications have reported on the usefulness of LLMs for a range of
linguistic annotation tasks [e.g. @yu2024; @heseltine2024] **maybe add
more?**
<!--# ELF: Yes, and specifying the kinds of linguistic tasks. E.g. on metaphor detection (Fuoli et al., 2025), annotation of speech acts (Yu et al. 2024; Su & Ye 2025), word sense or function disambiguation (Fonteyn et al., 2023; Morin & Marttinen Larsson 2025). + mostly on English so far -->.

While LLMs are increasingly prevalent in both linguistic research and
our everyday lives, they also raise a number of concerns. For research
purposes, specifically, their 'black-box' nature raises concerns over
the transparency, reproducibility, replicability of the research
produced with their assistance. Note, that although the terms have been
used interchangeably in the past, we distinguish between reproducibility
and replicability following the definitions of the FORRT
[@parsonsCommunitysourcedGlossaryOpen2022] and Turing Way
[@theturingwaycommunityTuringWayHandbook2022] communities. We thus
define *reproducibility* as the ability of an independent researcher or
team to obtain the same results as an original study using the original
data and methods (see @fig-ReproducibilityTerminology). This is in
contrast to *replicability*, where the same methods are applied, but on
different data; and robustness, where the same data, but different
methods are used. Only if an effect can be reliably observed across
different datasets with different methods, can we claim that it is
*generalisable*.

![Term disambiguation [@theturingwaycommunityTuringWayHandbook2022, CC
BY
4.0]](figures/TuringWay_reproducible-matrix.jpg){#fig-ReproducibilityTerminology
out-width="70%"
fig-alt="A table defining four terms: reproducible, replicable, robust, and generalisable. For reproducibility, data and analysis are the same. For replicability, data is different and the analysis is the same. For robustness, data is the same and the analysis is different. For generalisability, data and analysis are both different."
width="60%"}

In this study, we seek to replicate @morin2025's methodological study,
whose title claims to present a "replicable method" for leveraging LLMs
for linguistic annotation. The original study was conducted using a
commercial, closed-source LLM: Anthropic's model Claude 3.5 Sonnet
[@morin2025: 4], which is no longer available for use. We therefore
first sought to apply the original methodological pipeline to the same
data, but using two locally-run
<!--# HB: is this accurate, since I used them via KI:connect API? -->,
open-source LLMs, whose parameters can be controlled to generate
reproducible results. In an extended replication, we subsequently re-ran
the task with a much simpler pipeline than that proposed in the original
paper. Finally, we repeated the classification task using a simpler and
widely used NLP tool from the BERT family, distilBERT.

While large, cloud-based, proprietary LLMs continue to top performance
benchmarks, open-source LLMs have made significant progress in closing
this performance gap [@manchanda: 4]. We consider open models to form a
crucial part of the open research ecosystem (alongside open materials,
open data, open code, open documentation, open access, etc.). Openness
promotes transparency, reproducibility, and quality control; which are
necessary features to support accessible, reliable, and credible
empirical findings and build trustworthy AI [@liesenfeld2023]. We also
believe that smaller, locally-run LLMs are worth considering for
environmental reasons: Training and operating LLMs with billions of
parameters requires substantial computational resources, including large
amounts of water (e.g. for cooling data centers) and electricity,
generating substantial carbon emissions
<!--# ELF: insert refs! HB: also cited from jegham2015 - or meaning more refs?-->.
Although AI becomes cheaper and faster, its global adoption is rapidly
increasing resource consumption [@jegham2025].

# Research aims and questions

This study aims to replicate the analysis published in:

> Morin, Cameron, and Matti Marttinen Larsson. 2025. “Large Corpora and
> Large Language Models: A Replicable Method for Automating Grammatical
> Annotation.” *Linguistics Vanguard* 11 (1): 501–10.
> <https://doi/10.1515/lingvan-2024-0228>.

The data associated with this project is publicly available in an OSF
repository [@morin2024]. It contains a "conversation transcript", from
which the iterative training process could be traced, and several
datasets for training and testing an LLM on the task of differentiating
between evaluative and non-evaluative uses of the verb *consider*. The
model outputs were evaluated against a human gold standard provided by
the authors.

We want to investigate the following research questions (RQs):

1.  Can smaller, open-source models perform as well or better on the
    same task as Claude Sonnet 3.5 based on the accuracy metrics
    reported in @morin2025?
2.  Can 1.
    <!--# HB: I did not find an acceptable formulation including, task, accuracy metrics etc., but I don't know if this explicit enough? -->
    be achieved with a significantly less extensive training process
    than in @morin2025?
3.  Can a small encoder-only transformer model from the BERT family,
    distilBERT, perform as well or better on the same task as Claude
    Sonnet 3.5 based on the accuracy metrics reported in @morin2025?

While the first RQ is examined in the close replication, the extended
replication is concerned with RQ2 and RQ3. The open-source models used
for this replication are Mistral Small 3.2 24B Instruct 2506
[@mistralai2026] (further referred to as "Mistral"), an open-weight
model released in 2025 under Apache 2.0 license by the French AI company
Mistral AI, and OpenAI GPT OSS 120B [@openai2026], an open-weight large
language model that was also released under Apache 2.0 license by OpenAI
(referred to as "GPT").

The results from this replication will be compared and assessed based on
the metrics @morin2025 provided in their evaluation: Accuracy,
Precision, Recall, F1 score, and Matthew's correlation coefficient
[@morin2025: 6]. To make our results measurable, we formulate the
following hypotheses:

Hypothesis 1.1: The model Mistral Small 3.2 24B Instruct 2506performs as
well as or better than Claude Sonnet 3.5 on the same task, as measured
by the accuracy metrics reported in @morin2025.

Hypothesis 1.2: The model OpenAI GPT OSS 120B performs as well as or
better than Claude Sonnet 3.5 on the same task, as measured by the
accuracy metrics reported in @morin2025.

Hypothesis 2: The task can be performed with a significantly less
extensive training process than that used by @morin2025, while achieving
performance comparable to or better than Claude Sonnet 3.5 according to
the accuracy metrics reported in @morin2025.

Hypothesis 3: DistilBERT, a small encoder-only transformer model from
the BERT family, performs as well as or better than Claude Sonnet 3.5 on
the same task, as measured by the accuracy metrics reported in
@morin2025.

<!--# HB: I dont know if this is too much or too redundant, but I now included both RQs and hypotheses and also split the first RQ into 2 hypotheses so we can give clearer answers. -->

# Close replication

@morin2025 present a methodological pipeline leveraging an LLM for the
automatic annotation of grammatical data, based on the case study of the
evaluative *consider* construction in English. Their annotation pipeline
consisted of a complex "conversation" with the model over several
iterative phases of training and testing. Using Anthropic's model Claude
3.5 Sonnet [@morin2025: 4], they were able to achieve an accuracy of 93%
compared to the gold standard provided by a human annotator [@morin2025:
6]. In a first replication, we aimed to repeat @morin2025's "replicable
method" as closely as possible, using the same data and matching the
training and testing to a conversation transcript the authors provided
[@morin2024], albeit with two open LLMs: Mistral Small 3.2 24B Instruct
[@mistralai2026] and OpenAI GPT OSS 120B [@openai2026].

```{r, echo=FALSE}
kable(
  metrics_table,
  col.names = c(
    "Metric",
    "Mistral",
    "GPT",
    "Claude (original study)"
  ),
  caption = "Classification metrics of the three models compared to the human gold standard"
)
```

Based on accuracy compared to the human gold standard, GPT showed
comparable and even slightly better performance than Claude in the
original study (GPT: `r round(100 * gpt_accuracy)`% vs. Claude
`r round(100 * claude_accuracy)`%), while Mistral performed considerably
worse (`r round(100 * mistral_accuracy)`%). This observation is
reflected in the other metrics as well: While GPT's performance is
similar or better in all evaluated metrics, Mistral fails to achieve
comparable results. While Mistral's 67% accuracy might initially sound
moderately good, investigating these results as 2 x 2 matrices **(see
`Appendix.R`**) reveals that it is heavily biased toward predicting
'non-evaluative'. This shows especially in the particularly low Recall
of
`r sprintf("%.1f%%", as.numeric(conf_matrix_mistral$byClass["Sensitivity"]) * 100)`,
since this metric answers the question how many positives of the actual
positives were correctly identified.

Looking at this result, it is important to consider the distribution of
*consider*-meanings in the testing dataset. The 'non-evaluative'
classification was the majority class
(`r majority_count`/`r test_total`,
`r round(100 * majority_accuracy)`%), and a classifier that always
predicted the majority class would have achieved an accuracy of
`r round(100 * majority_accuracy)`%. The imbalance in this dataset makes
it especially important to consider other metrics when assessing model
performance.

**Inter-rater agreement metrics support these observations: GPT showed
strong agreement with the human gold standard (**$\kappa$ **=
`r round(gpt_kappa, 2)`), whereas Mistral showed only slight agreement
beyond chance (**$\kappa$ **= `r round(mistral_kappa, 2)`).**
<!--# HB: should I drop the inter-rater agreement part now since we are looking at so many other metrics? -->

The results suggest that the (close) replicability of the approach
evidently depends on the comparability of the models. Recalling RQ1, the
GPT performed slightly better than Claude 3.5 Sonnet, whereas the
smaller Mistral model did not. Therefore, Hypothesis 1.1 is rejected,
while hypothesis 1.2 is supported by the results. On a methodological
level, we found that essential information to independently replicate
the study (e.g., whether the web interface or an API was used, analysis
code, etc.) was missing - forcing us to make some educated guesses as to
the exact procedure.

# Extended replication

The close replication raised questions about both the necessity of such
an extensive training and testing procedure, and the need for large
commercial LLMs to complete this kind of linguistic task. This led us to
extend our replication to address both of these issues. First, we tested
the need for the extensive 'in-conversation' training process outlined
in @morin2025. To this end, we applied a much simpler few-shot training
approach with the best-performing open LLM (GPT) from our first
replication using the same data.

```{r, echo=FALSE}
kable(
  metrics_table_ext,
  col.names = c(
    "Metric",
    "GPT (close replication)",
    "GPT (reduced training)",
    "Claude (original study)"
  ),
  caption = "Classification metrics of both GPT analyses and Claude compared to the human gold standard"
)
```

When only using the pretraining set and directly proceeding to testing
afterwards, GPT achieved an accuracy of
`r round(100 * gpt_reduced_2_accuracy)`%
(`r gpt_reduced_2_agreement`/`r gpt_reduced_2_total`), with a Cohen's
$\kappa$ of `r round(gpt_reduced_2_kappa, 2)`.
<!--# HB: leave out Cohens kappa??) --> While the accuracy is equivalent
to the GPT result in the close replication, the other metrics reveal
some minor differences in classification performance: Recall is 100%,
meaning that GPT did not miss any positives ('evaluative' sentences) in
the extended replication. Compared to the original study, the accuracy
metrics are comparable: Precision is slightly lower (GPT:
`r round(conf_matrix_gpt_reduced_2$byClass["Pos Pred Value"] * 100, 1)`%
vs. Claude:
`r round(conf_matrix_claude$byClass["Pos Pred Value"] * 100, 1)`%),
meaning GPT predicted more false positives ('evaluative' for actual
'non-evaluative'). All the other metrics are higher than Claude for the
GPT replication with fewer training. Therefore, these results provide
support for hypothesis 2.
<!--# HB: this still depends on how exactly we handle the metrics/benchmarks in the hypotheses.-->

```{r, echo=FALSE}
kable(
  metrics_table_bert,
  col.names = c(
    "Metric",
    "DistilBERT (pretraining only)",
    "DistilBERT (extended training)",
    "Claude (original study)"
  ),
  caption = "Classification metrics of distilBERT and CLaude compared to the human gold standard"
)
```

As a second extension, we replicated the annotation study using a
fine-tuned distilBERT model [@sanh]. This encoder-only transformer is
considerably smaller than even a small generative LLM. The first run was
similar to the reduced training approach with GPT: only the pretraining
set was used for training the model, and, after using the "unsupervised
training" sets for validation, we proceeded to testing. DistilBERT
achieved an accuracy of `r round(100 * bert_accuracy)`%
(`r bert_agreement`/`r bert_total`), matching the majority-class
baseline, as it classified all test items as 'non-evaluative'. In a
second round, we combined the two unsupervised training sets and the
pretraining set into one bigger training dataset, and ran the same test
(see `BERT_extended_script.R`). In this case, distilBERT categorized one
sentence as 'evaluative', and is still almost completely aligning with
the majority baseline. For the first analysis with distilBERT, most of
the metrics become meaningless when there are no positive predictions at
all. Recall is
`r round(conf_matrix_bert$byClass["Sensitivity"] * 100, 1)`%, because
there are 39 actual 'evaluative' cases, but the model correctly
identifies 0 of them. For the second round with the expanded training
dataset, numbers could be defined because there was a positive
prediction. While most of them are very low compared to Claude, the
reported Precision is
`r round(conf_matrix_bert_extended$byClass["Pos Pred Value"] * 100, 1)`%.
This metric measures how often the model is correct when it predicts
positives. Since the model only predicted 'evaluative' in one singular
case, this result is misleading. Based on both the accuracy and the
further examined metrics for distilBERT, hypothesis 3 is rejected.

In summary, our results suggest that, when using a comparable
open-source LLM, the training process need not be as elaborate as in
@morin2025 to accomplish similarly high accuracy rates. However, the
smaller Mistral model and the distilBERT model showed a strong tendency
toward the majority class and performed only slightly above or exactly
at the majority-class baseline, indicating to not be suitable to
successfully complete this kind of linguistic task.

# Discussion

## Suitability of LLMs and training process

As the results of this replication indicate, an open-source LLM can
compete with a closed LLM on this linguistic classification task. As
already mentioned in the introduction, open models constitute an
important part of open research. Aside from limited availability,
closed, commercial LLMs are often developed intransparently and without
ethical oversight [@liesenfeld2023]. Since their training relies on vast
amounts of data, including text from the internet, publicly available
datasets, or proprietary sources, they are susceptible to passive
privacy leakage, whereby sensitive information contained in training
data may be revealed. Also, LLMs can be vulnerable to different kinds of
active privacy attacks [@yan2025: 1].

## "Replicable" methods

Method-focused papers with (commercial) LLMs -\> what does that mean?
"replicable" pipeline (!) -\> make things replicable importance of
reproducibility **(might turn into 2 chapters?)**

missing: report of test set distribution, code, programming language,
information, also human gold standard was only from 1 person. There is
no information of if inter-rater agreemtn was tested with this

# Conclusion (whole paper: max. 4000 words)

While LLMs have shown promising results in a range of tasks, further
exploration is needed to effectively and reliably leverage them for
linguistic data annotation [@lidiiaostyakova2025: 131].

*what we may learn from the extensions about the approach and LLM use in
corpus studies method and replicability open science practices esp.
important when we throw LLMs on everything*
