---
title: "LLMs for corpus annotation: Close and extended replications of Morin & Marttinen Larsson (2025) using smaller open LLMs and BERT"
author: Hannah Boxleitner & Elen Le Foll
format: html
bibliography: references.bib
csl: unified-style-sheet-for-linguistics.csl
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| label: close_replication_results
#| include: false

library(readr)
library(irr)

comparison_all <- read_csv2(
  "replication_data/comparison_all_results.csv",
  show_col_types = FALSE
)

# Test distribution

test_distribution <- table(
  comparison_all$Human_classification
)

majority_count <- max(test_distribution)
test_total <- nrow(comparison_all)
majority_accuracy <- majority_count / test_total

# Accuracy
gpt_accuracy <- mean(
  comparison_all$Human_classification == comparison_all$GPT_classification
)

mistral_accuracy <- mean(
  comparison_all$Human_classification == comparison_all$Mistral_classification
)

claude_accuracy <- mean(
  comparison_all$Human_classification == comparison_all$Claude_classification
)

# Cohen's kappa
gpt_kappa <- kappa2(
  comparison_all[c("Human_classification", "GPT_classification")]
)$value

mistral_kappa <- kappa2(
  comparison_all[c("Human_classification", "Mistral_classification")]
)$value

claude_kappa <- kappa2(
  comparison_all[c("Human_classification", "Claude_classification")]
)$value


```

```{r}
#| label: reduced-training-2
#| include: false

library(readr)
library(irr)

gpt_reduced_2 <- read_csv2(
  "reduced_training/GPT_reduced_training_2_results.csv",
  show_col_types = FALSE
)

# Accuracy
gpt_reduced_2_accuracy <- mean(
  gpt_reduced_2$Human_classification ==
    gpt_reduced_2$GPT_classification
)

# Number of agreements
gpt_reduced_2_agreement <- sum(
  gpt_reduced_2$Human_classification ==
    gpt_reduced_2$GPT_classification
)

# Cohen's kappa
gpt_reduced_2_kappa <- kappa2(
  gpt_reduced_2[
    c("Human_classification", "GPT_classification")
  ]
)$value

# Number of test items
gpt_reduced_2_total <- nrow(gpt_reduced_2)

```

```{r}
#| label: bert-results
#| include: false

load("BERT_transcripts/BERT_experiment_workspace.RData")

bert_accuracy <- mean(
  preds_test == labels_test
)

bert_agreement <- sum(
  preds_test == labels_test
)

bert_total <- length(labels_test)
```

```{r}
#| label: bert-extended-results
#| include: false

library(knitr)

load("BERT_extended_transcripts/BERT_extended.RData")

bert_extended_accuracy <- mean(
  preds_test_extended == labels_test_extended
)

bert_extended_agreement <- sum(
  preds_test_extended == labels_test_extended
)

bert_extended_total <- length(labels_test_extended)
```

# Abstract

Large Language Models (LLMs) are increasingly being leveraged for the
annotation of large corpora. In an early evaluation of such an
application, @morin2025 presented "a replicable method for automating
grammatical annotation" using Anthropic's model Claude 3.5 Sonnet
[@morin2025: 4]. Their annotation pipeline consisted of a complex
"conversation" with the model over several iterative phases of training
and testing. They were able to achieve an accuracy of 93% compared to
the gold standard provided by a human annotator [@morin2025: 6] and
conclude that the pipeline may also be effective for other linguistic
tasks in other languages [@morin2025: 7].

We aimed to repeat @morin2025's "replicable method" as closely as
possible, using the same data, albeit with two open LLMs, Mistral Small
3.2 24B Instruct [@mistralai2026] and OpenAI GPT OSS 120B [@openai2026].
While the smaller Mistral model did not perform as well on the task
(accuracy of 67%), the open GPT model yielded very similar results
(accuracy of 94%). Although the authors published their data and a full
transcript of their "conversation" with the model [@morin2024], we found
that essential information to independently replicate the study (e.g.,
whether the web interface or an API was used, analysis code, etc.) was
missing - forcing us to make some educated guesses as to the exact
procedure. Our results suggest that, as of early 2026, large commercial
LLMs may not be necessary to complete this kind of linguistic task in
contemporary English.

Next, we tested the need for the extensive 'in-conversation' training
process outlined in @morin2025. To this end, we conducted a second,
extended replication applying a much simpler few-shot training approach
with the best-performing open LLM from our first replication. Our
results suggest that the training process need not be as elaborate as in
@morin2025 to accomplish similarly high accuracy rates.

Finally, we replicated the annotation study using a fine-tuned
DistilBERT model [@sanh]. Although this encoder-only transformer is
considerably smaller than even a small generative LLM, it performed
approximately on par with the 24B Mistral model used in our first
replication. These findings indicate that task-specific supervised
models may provide a competitive and certainly more computationally
efficient alternative to generative LLMs.

Our results have important implications for the replicability of
corpus-linguistic research involving LLMs as part of the annotation
process. We argue that, when choosing a method, issues around model
accessibility, data leakage, and carbon footprint can and should be
taken into consideration.

# Introduction

Over the past decades, linguists have gained access to ever larger
corpora of natural language. This poses great opportunities, but also
considerably increases the workload associated with the manual and/or
(semi-)automated preprocessing and annotation of large amounts of data.
The rise of large language models (LLMs) is therefore an exciting
development for linguistics research, offering new possibilities for
complementing and extending existing automated methods for processing
larger datasets and corpora. In recent years, a growing number of
publications have reported on the usefulness of LLMs for a range of
linguistic annotation tasks [e.g. @yu2024; @heseltine2024] **maybe add
more?**
<!--# ELF: Yes, and specifying the kinds of linguistic tasks. E.g. on metaphor detection (Fuoli et al., 2025), annotation of speech acts (Yu et al. 2024; Su & Ye 2025), word sense or function disambiguation (Fonteyn et al., 2023; Morin & Marttinen Larsson 2025). + mostly on English so far -->.

While LLMs are increasingly prevalent in both linguistic research and
our everyday lives, they also raise a number of concerns. For research
purposes, specifically, their 'black-box' nature raises concerns over
the transparency, reproducibility, replicability of the research
produced with their assistance. Note, that although the terms have been
used interchangeably in the past, we distinguish between reproducibility
and replicability following the definitions of the FORRT
[@parsonsCommunitysourcedGlossaryOpen2022] and Turing Way
[@theturingwaycommunityTuringWayHandbook2022] communities. We thus
define *reproducibility* as the ability of an independent researcher or
team to obtain the same results as an original study using the original
data and methods (see @fig-ReproducibilityTerminology). This is in
contrast to *replicability*, where the same methods are applied, but on
different data; and robustness, where the same data, but different
methods are used. Only if an effect can be reliably observed across
different datasets with different methods, can we claim that it is
*generalisable*.

![Term disambiguation [@theturingwaycommunityTuringWayHandbook2022, CC
BY
4.0]](figures/TuringWay_reproducible-matrix.jpg){#fig-ReproducibilityTerminology
out-width="70%"
fig-alt="A table defining four terms: reproducible, replicable, robust, and generalisable. For reproducibility, data and analysis are the same. For replicability, data is different and the analysis is the same. For robustness, data is the same and the analysis is different. For generalisability, data and analysis are both different."
width="60%"}

In this study, we seek to replicate @morin2025's methodological study,
whose title claims to present a "replicable method" for leveraging LLMs
for linguistic annotation. The original study was conducted using a
commercial, closed-source LLM: Anthropic's model Claude 3.5 Sonnet
[@morin2025: 4], which is no longer available for use. We therefore
first sought to apply the original methodological pipeline to the same
data, but using two locally-run
<!--# HB: is this accurate, since I used them via KI:connect API? -->,
open-source LLMs, whose parameters can be controlled to generate
reproducible results. In an extended replication, we subsequently re-ran
the task with a much simpler pipeline than that proposed in the original
paper. Finally, we repeated the classification task using a simpler and
widely used NLP tool from the BERT family, distilBERT.

While large, cloud-based, proprietary LLMs continue to top performance
benchmarks, open-source LLMs have made significant progress in closing
this performance gap [@manchanda: 4]. We consider open models to form a
crucial part of the open research ecosystem (alongside open materials,
open data, open code, open documentation, open access, etc.). Openness
promotes transparency, reproducibility, and quality control; which are
necessary features to support accessible, reliable, and credible
empirical findings and build trustworthy AI [@liesenfeld2023]. We also
believe that smaller, locally-run LLMs are worth considering for
environmental reasons: Training and operating LLMs with billions of
parameters requires substantial computational resources, including large
amounts of water (e.g. for cooling data centers) and electricity,
generating substantial carbon emissions
<!--# ELF: insert refs! HB: also cited from jegham2015 - or meaning more refs?-->.
Although AI becomes cheaper and faster, its global adoption is rapidly
increasing resource consumption [@jegham2025].

# Research aims and questions

This study aims to replicate the analysis published in:

> Morin, Cameron, and Matti Marttinen Larsson. 2025. “Large Corpora and
> Large Language Models: A Replicable Method for Automating Grammatical
> Annotation.” *Linguistics Vanguard* 11 (1): 501–10.
> <https://doi/10.1515/lingvan-2024-0228>.

The data associated with this project is publicly available in an OSF
repository [@morin2024]. It contains a "conversation transcript", from
which the iterative training process could be traced, and several
datasets for training and testing an LLM on the task of differentiating
between evaluative and non-evaluative uses of the verb *consider*. The
model outputs were evaluated against a human gold standard provided by
the authors.

We want to investigate the following research questions (RQs):

1.  Can smaller, open-source models perform as well or better on the
    same task as Claude Sonnet 3.5 based on the accuracy metrics
    reported in @morin2025?
2.  Can 1.
    <!--# HB: I did not find an acceptable formulation including, task, accuracy metrics etc., but this is not explicit enough I guess? -->
    be achieved with a significantly less extensive training process
    than in @morin2025?
3.  Can a small encoder-only transformer model from the BERT family,
    distilBERT, perform as well or better on the same task as Claude
    Sonnet 3.5 based on the accuracy metrics reported in @morin2025?

The open-source models used for this replication are Mistral Small 3.2
24B Instruct 2506 [@mistralai2026] (further referred to as "Mistral"),
an open-weight model released in 2025 under Apache 2.0 license by the
French AI company Mistral AI, andOpenAI GPT OSS 120B [@openai2026], an
open-weight large language model that was also released under Apache 2.0
license by OpenAI (referred to as "GPT").

The results from this replication will be compared and assessed based on
the metrics @morin2025 provided in their evaluation: Accuracy,
Precision, Recall, F1 score, and Matthew's correlation coefficient
[@morin2025: 6].

<!--# HB: I was unsure if we should formulate these questions into hypotheses as well, because it would be somewhat redundant, but also contribute to make it clearly testable? -->

While the first RQ is examined in the close replication, the extended
replication is concerned with RQ2 and RQ3.

<!--# HB: Should we get into more detail here e.g. how the close replication has qualities of both reproduction and replication? Or is this too much?-->

# Close replication

@morin2025 present a methodological pipeline leveraging an LLM for the
automatic annotation of grammatical data, based on the case study of the
evaluative *consider* construction in English. Their annotation pipeline
consisted of a complex "conversation" with the model over several
iterative phases of training and testing. Using Anthropic's model Claude
3.5 Sonnet [@morin2025: 4], they were able to achieve an accuracy of 93%
compared to the gold standard provided by a human annotator [@morin2025:
6]. In a first replication, we aimed to repeat @morin2025's "replicable
method" as closely as possible, using the same data and matching the
training and testing to a conversation transcript the authors provided
[@morin2024], albeit with two open LLMs: Mistral Small 3.2 24B Instruct
[@mistralai2026] and OpenAI GPT OSS 120B [@openai2026].

Based on accuracy compared to the human gold standard, GPT showed
comparable and even slightly better performance than Claude in the
original study (GPT: `r round(100 * gpt_accuracy)`% vs. Claude
`r round(100 * claude_accuracy)`%), while Mistral performed considerably
worse (`r round(100 * mistral_accuracy)`%). Inter-rater agreement
metrics support these observations: GPT showed strong agreement with the
human gold standard ($\kappa$\$\\kappa\$\$\\kappa\$\$\\kappa\$ =
`r round(gpt_kappa, 2)`), whereas Mistral showed only slight agreement
beyond chance ($\kappa$ = `r round(mistral_kappa, 2)`). While Mistral's
67% accuracy might initially sound moderately good, investigating these
results as 2 x 2 matrices **(see Appendix.R**) reveals that it is
heavily biased toward predicting 'non-evaluative'. Looking at the
distribution of *consider*-meanings in the testing dataset, the
'non-evaluative' classification was the majority class
(`r majority_count`/`r test_total`,
`r round(100 * majority_accuracy)`%), and a classifier that always
predicted the majority class would achieve an accuracy of
`r round(100 * majority_accuracy)`%.

The results suggest that the (close) replicability of the approach
evidently depends on the comparability of the models. On a
methodological level, we found that essential information to
independently replicate the study (e.g., whether the web interface or an
API was used, analysis code, etc.) was missing - forcing us to make some
educated guesses as to the exact procedure.

# Extended replication

The close replication raised questions about both the necessity of such
an extensive training and testing procedure, and the need for large
commercial LLMs to complete this kind of linguistic task. This led us to
extend our replication to address both of these issues. First, we tested
the need for the extensive 'in-conversation' training process outlined
in @morin2025. To this end, we applied a much simpler few-shot training
approach with the best-performing open LLM (GPT) from our first
replication using the same data. When only using the pretraining set and
directly proceeding to testing afterwards, GPT achieved an accuracy of
`r round(100 * gpt_reduced_2_accuracy)`%
(`r gpt_reduced_2_agreement`/`r gpt_reduced_2_total`), with a Cohen's
$\kappa$ of `r round(gpt_reduced_2_kappa, 2)`.

As a second extension, we replicated the annotation study using a
fine-tuned DistilBERT model [@sanh]. This encoder-only transformer is
considerably smaller than even a small generative LLM. Again, only the
pretraining set was used for training the model, and, after using the
"unsupervised training" sets for validation, we proceeded to testing.
BERT achieved an accuracy of `r round(100 * bert_accuracy)`%
(`r bert_agreement`/`r bert_total`), matching the majority-class
baseline, as it classified all test items as 'non-evaluative'. In a
second round, we included the two unsupervised training and the
pretraining set into one bigger dataset, and ran the same test (see
`BERT_extended_script.R`). In this case, distilBERT actually categorized
one sentence as 'evaluative', and is still almost completely aligning
with the majority baseline.

These results suggest that, when using a comparable LLM, the training
process need not be as elaborate as in @morin2025 to accomplish
similarly high accuracy rates. However, the distilBERT and the smaller
Mistral model showed a strong tendency toward the majority class and
performed only slightly above or exactly at the majority-class baseline,
indicating to not be suitable to successfully complete this kind of
linguistic task.

# Discussion

## Suitability of LLMs and training process

While with a

Closed, commercial LLMs are often developed intransparently and without
ethical oversight [@liesenfeld2023]. Since their training relies on vast
amounts of data, including text from the internet, publicly available
datasets, or proprietary sources, they are susceptible to passive
privacy leakage, whereby sensitive information contained in training
data may be revealed. Also, LLMs can be vulnerable to different kinds of
active privacy attacks [@yan2025: 1].

## "Replicable method"

Method-focused papers with (commercial) LLMs -\> what does that mean?
"replicable" pipeline (!) -\> make things replicable importance of
reproducibility **(might turn into 2 chapters?)**

# Conclusion (whole paper: max. 4000 words)

While LLMs have shown promising results in a range of tasks, further
exploration is needed to effectively and reliably leverage them for
linguistic data annotation [@lidiiaostyakova2025: 131].

*what we may learn from the extensions about the approach and LLM use in
corpus studies method and replicability open science practices esp.
important when we throw LLMs on everything*
