---
title: "Appendix"
subtitle: "LLMs for corpus annotation: Close and extended replications of Morin & Marttinen Larsson (2025) using smaller open LLMs"
date: last-modified
format: pdf
editor: visual
execute:
  echo: false
---

# Full results

```{r}
#| label: setup
#| include: false
#| warning: false
#| message: false

library(readr)
library(irr)
library(knitr)
library(caret)
library(kableExtra)
```

```{r}
#| label: appendix-results-close-replication
#| include: false

# everything needed for test distribution, close replication, and later results
comparison_all <- read_csv2(
  "replication_data/comparison_all_results.csv",
  show_col_types = FALSE
)

# Test set distribution
test_distribution <- table(
  comparison_all$Human_classification
)

majority_count <- max(test_distribution)
test_total <- nrow(comparison_all)
majority_accuracy <- majority_count / test_total

# Accuracy
gpt_accuracy <- mean(
  comparison_all$Human_classification ==
    comparison_all$GPT_classification
)

mistral_accuracy <- mean(
  comparison_all$Human_classification ==
    comparison_all$Mistral_classification
)

claude_accuracy <- mean(
  comparison_all$Human_classification ==
    comparison_all$Claude_classification
)

# Number of agreements
gpt_agreement <- sum(
  comparison_all$Human_classification ==
    comparison_all$GPT_classification
)

mistral_agreement <- sum(
  comparison_all$Human_classification ==
    comparison_all$Mistral_classification
)

claude_agreement <- sum(
  comparison_all$Human_classification ==
    comparison_all$Claude_classification
)

# Classification order for confusion matrices
classification_levels <- c(
  "evaluative",
  "non-evaluative"
)
```

## Distribution of classifications in the testing set provided by Morin & Marttinen Larsson (2025)

```{r}
#| label: tbl-test-distribution
#| tbl-cap: "Distribution of classifications in the testing set according to the human gold standard"

test_distribution_table <- data.frame(
  Classification = names(test_distribution),
  n = as.vector(test_distribution),
  Percentage = paste0(
    round(100 * as.vector(test_distribution) / test_total),
    "%"
  )
)

test_distribution_table <- rbind(
  test_distribution_table,
  data.frame(
    Classification = "Total",
    n = test_total,
    Percentage = "100%"
  )
)

kable(test_distribution_table)
```

## Close replication

The following tables represent the results of the close replication.

### Accuracy of all models compared to the human gold standard

```{r}
#| label: close-rep-tbl-full-results
#| tbl-cap: "Classification accuracy and agreement with the human gold standard"

results_table <- data.frame(
  Model = c("Claude", "GPT", "Mistral"),
  Accuracy = c(
    paste0(round(100 * claude_accuracy), "%"),
    paste0(round(100 * gpt_accuracy), "%"),
    paste0(round(100 * mistral_accuracy), "%")
  ),
  Alignments = c(
    paste0(claude_agreement, "/", nrow(comparison_all)),
    paste0(gpt_agreement, "/", nrow(comparison_all)),
    paste0(mistral_agreement, "/", nrow(comparison_all))
  )
  )

kable(results_table)

```

### Accuracy as 2x2 matrices

```{r}
#| label: tbl-gpt-close-confusion
#| tbl-cap: "GPT classifications compared with the human gold standard"
#| warning: false
#| message: false

# Reference
actual_GPT_close <- factor(
  comparison_all$Human_classification,
  levels = classification_levels
)

# Prediction
predicted_GPT_close <- factor(
  comparison_all$GPT_classification,
  levels = classification_levels
)

# Confusion matrix
conf_matrix_GPT_close <- confusionMatrix(
  data = predicted_GPT_close,
  reference = actual_GPT_close,
  positive = "evaluative"
)

#print(conf_matrix_GPT_close$table)
#kable(conf_matrix_GPT_close$table)

# Create table output with model names
kable(
  conf_matrix_GPT_close$table,
  caption = "Prediction (GPT) vs Reference (Human)")

```

```{r}
#| label: tbl-mistral-confusion
#| tbl-cap: "Mistral classifications compared with the human gold standard"
#| warning: false
#| message: false

actual_mistral <- factor(
  comparison_all$Human_classification,
  levels = classification_levels
)

predicted_mistral <- factor(
  comparison_all$Mistral_classification,
  levels = classification_levels
)

conf_matrix_mistral <- confusionMatrix(
  data = predicted_mistral,
  reference = actual_mistral,
  positive = "evaluative"
)

kable(
  conf_matrix_mistral$table,
  caption = "Prediction (Mistral) vs Reference (Human)")
```

```{r}
#| label: tbl-claude-confusion
#| tbl-cap: "Claude classifications compared with the human gold standard"
#| warning: false
#| message: false

actual_claude <- factor(
  comparison_all$Human_classification,
  levels = classification_levels
)

predicted_claude <- factor(
  comparison_all$Claude_classification,
  levels = classification_levels
)

conf_matrix_claude <- confusionMatrix(
  data = predicted_claude,
  reference = actual_claude,
  positive = "evaluative"
)

kable(
  conf_matrix_claude$table,
  caption = "Prediction (Claude) vs Reference (Human)")
```

### Accuracy metrics for all models in close replication

```{r}
#| label: tbl-accuracy-metrics-close
#| tbl-cap: "Classification performance of the three models compared with the human gold standard"
#| warning: false
#| message: false

# Calculate Matthews correlation coefficient
calculate_mcc <- function(cm) {
  
  # Extract the four cells from confusion matrix
  TP <- cm[1, 1]
  FN <- cm[2, 1]
  FP <- cm[1, 2]
  TN <- cm[2, 2]
  
  # Calculate denominator
  denominator <- sqrt(
    (TP + FP) *
      (TP + FN) *
      (TN + FP) *
      (TN + FN)
  )
  
  # Avoid division by 0
  if (denominator == 0) {
    return(NA_real_)
  }
  
  # Calculate Matthews correlation coefficient
  (TP * TN - FP * FN) / denominator
}


# Create table with metrics for all three models
metrics_table <- data.frame(
  
  # Names of metrics
  Metric = c(
    "Accuracy",
    "Precision",
    "Recall",
    "F1-score",
    "Matthews correlation coefficient"
  ),
  
  # Mistral metrics
  Mistral = c(
    as.numeric(conf_matrix_mistral$overall["Accuracy"]),
    as.numeric(conf_matrix_mistral$byClass["Pos Pred Value"]),
    as.numeric(conf_matrix_mistral$byClass["Sensitivity"]),
    as.numeric(conf_matrix_mistral$byClass["F1"]),
    calculate_mcc(conf_matrix_mistral$table)
  ),
  
  # GPT metrics
  GPT = c(
    as.numeric(conf_matrix_GPT_close$overall["Accuracy"]),
    as.numeric(conf_matrix_GPT_close$byClass["Pos Pred Value"]),
    as.numeric(conf_matrix_GPT_close$byClass["Sensitivity"]),
    as.numeric(conf_matrix_GPT_close$byClass["F1"]),
    calculate_mcc(conf_matrix_GPT_close$table)
  ),
  
  # Claude metrics
  Claude = c(
    as.numeric(conf_matrix_claude$overall["Accuracy"]),
    as.numeric(conf_matrix_claude$byClass["Pos Pred Value"]),
    as.numeric(conf_matrix_claude$byClass["Sensitivity"]),
    as.numeric(conf_matrix_claude$byClass["F1"]),
    calculate_mcc(conf_matrix_claude$table)
  )
)


# Format Accuracy as whole percentage
metrics_table[1, -1] <- paste0(
  round(as.numeric(metrics_table[1, -1]) * 100, 0),
  "%"
)


# Format Precision, Recall and F1-score (as percentages with one decimal place)
metrics_table[2:4, -1] <- lapply(
  metrics_table[2:4, -1],
  function(x) {
    paste0(
      round(as.numeric(x) * 100, 1),
      "%"
    )
  }
)


# Format MCC to two decimal places (not percent)
metrics_table[5, -1] <- lapply(
  metrics_table[5, -1],
  function(x) {
    sprintf("%.2f", as.numeric(x))
  }
)


# Final table
kable(
  metrics_table,
  col.names = c(
    "Metric",
    "Mistral",
    "GPT",
    "Claude (original study)"
  )
)
```

## Extended replication with GPT: simpler training process

### Accuracy compared to the human gold standard

```{r}
#| label: appendix-results-GPT-reduced-training-2-extended
#| include: false

gpt_reduced_2 <- read_csv2(
  "reduced_training/GPT_reduced_training_2_results.csv",
  show_col_types = FALSE
)

# Accuracy
gpt_reduced_2_accuracy <- mean(
  gpt_reduced_2$Human_classification ==
    gpt_reduced_2$GPT_classification
)

# Number of agreements
gpt_reduced_2_agreement <- sum(
  gpt_reduced_2$Human_classification ==
    gpt_reduced_2$GPT_classification
)


# Number of test items
gpt_reduced_2_total <- nrow(gpt_reduced_2)

```

```{r}
#| label: tbl-GPT-reduced-training-2
#| tbl-cap: "GPT performance in the reduced training replication (only pretraining dataset and testing)"

reduced_training_2_table <- data.frame(
  Accuracy = paste0(
    round(100 * gpt_reduced_2_accuracy),
    "%"
  ),
  Alignments = paste0(
    gpt_reduced_2_agreement,
    "/",
    gpt_reduced_2_total
  )
  )

kable(reduced_training_2_table)
```

### Accuracy as 2x2 matrices

```{r}
#| label: tbl-gpt-reduced-2-confusion
#| tbl-cap: "GPT classifications with few-shot training compared to the human gold standard"
#| warning: false
#| message: false

actual_gpt_reduced_2 <- factor(
  gpt_reduced_2$Human_classification,
  levels = classification_levels
)

predicted_gpt_reduced_2 <- factor(
  gpt_reduced_2$GPT_classification,
  levels = classification_levels
)

conf_matrix_gpt_reduced_2 <- confusionMatrix(
  data = predicted_gpt_reduced_2,
  reference = actual_gpt_reduced_2,
  positive = "evaluative"
)

kable(
  conf_matrix_gpt_reduced_2$table,
  caption = "Prediction (GPT reduced training) vs Reference (Human)"
)
```

### Accuracy metrics of GPT few-shot training extension compared to Claude in original study

```{r}
#| label: tbl-accuracy-metrics-reduced-2
#| tbl-cap: "Classification performance of GPT after reduced training and Claude compared with the human gold standard"
#| warning: false
#| message: false

# Calculate Matthews correlation coefficient
calculate_mcc <- function(cm) {
  
  # Extract the four cells from confusion matrix
  TP <- cm[1, 1]
  FN <- cm[2, 1]
  FP <- cm[1, 2]
  TN <- cm[2, 2]
  
  # Calculate denominator
  denominator <- sqrt(
    (TP + FP) *
      (TP + FN) *
      (TN + FP) *
      (TN + FN)
  )
  
  # Avoid division by 0
  if (denominator == 0) {
    return(NA_real_)
  }
  
  # Calculate Matthews correlation coefficient
  (TP * TN - FP * FN) / denominator
}


# Create table with metrics for both models
metrics_table <- data.frame(
  
  # Names of metrics
  Metric = c(
    "Accuracy",
    "Precision",
    "Recall",
    "F1-score",
    "Matthews correlation coefficient"
  ),
  
  # GPT reduced training metrics
  `GPT (reduced training)` = c(
    as.numeric(conf_matrix_gpt_reduced_2$overall["Accuracy"]),
    as.numeric(conf_matrix_gpt_reduced_2$byClass["Pos Pred Value"]),
    as.numeric(conf_matrix_gpt_reduced_2$byClass["Sensitivity"]),
    as.numeric(conf_matrix_gpt_reduced_2$byClass["F1"]),
    calculate_mcc(conf_matrix_gpt_reduced_2$table)
  ),
  
  # Claude metrics
  Claude = c(
    as.numeric(conf_matrix_claude$overall["Accuracy"]),
    as.numeric(conf_matrix_claude$byClass["Pos Pred Value"]),
    as.numeric(conf_matrix_claude$byClass["Sensitivity"]),
    as.numeric(conf_matrix_claude$byClass["F1"]),
    calculate_mcc(conf_matrix_claude$table)
  )
)


# Format Accuracy as whole percentage
metrics_table[1, -1] <- paste0(
  round(as.numeric(metrics_table[1, -1]) * 100, 0),
  "%"
)


# Format Precision, Recall and F1-score (as percentages with one decimal place)
metrics_table[2:4, -1] <- lapply(
  metrics_table[2:4, -1],
  function(x) {
    paste0(
      round(as.numeric(x) * 100, 1),
      "%"
    )
  }
)


# Format MCC to two decimal places
metrics_table[5, -1] <- lapply(
  metrics_table[5, -1],
  function(x) {
    sprintf("%.2f", as.numeric(x))
  }
)


# Final table
kable(
  metrics_table,
  col.names = c(
    "Metric",
    "GPT (reduced training)",
    "Claude (original study)"
  )
)
```

# Further details on the study

## Technical details regarding the models

The open LLMs were accessed remotely via an API [@ki:conne]. The generation parameters in the API call were set to *temperature* = 0, *top~p~* = 1, and a constant random seed was used. Setting *temperature* to 0, meaning the model always chooses the most probable next token, removes random sampling and therefore makes additional constraints such as *top~k~* (sampling only from the k most probable tokens) unnecessary and potentially confounding. Even though these measures were taken to improve reproducibility, identical outputs across runs cannot be guaranteed.

## Data leakage

For this study, it is assumed that the models, functioning as several "annotators", independently label the data without knowledge of each other's work. It can only be *assumed* because there is a possibility that the models might have the gold standard provided in the original study as part of their training data, which would mean they would not be totally independent. The Mistral model used for this study was released in June 2025 [@mistralai2026], and the GPT model in August 2025 [@openai2026]. @morin2025, who used data from the NOW corpus and Sketch Engine’s EnTenTen21 corpus for their study [@morin2025: 3], created their OSF-database containing the (annotated) datasets in November 2024 [@morin2024]. When and on which data the models were trained exactly remains unclear, but since they were released after the repository went public, the possibility of data leakage cannot be entirely ruled out.
